---
title: "Complete Reference: estimate_concentration_field()"
subtitle: "Physics, solvers, and practical tuning"
author: "Raredon Laboratory"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Reference: estimate_concentration_field()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)
```

---

# Overview

Spatial transcriptomics data gives us discrete transcript or cell locations, but many biological
processes — ligand gradients, cytokine fields, metabolite concentrations — are continuous.
`estimate_concentration_field()` bridges this gap by computing the **steady-state concentration
field** that a set of point sources (mRNA transcripts, detected proteins) would produce under a
physically motivated diffusion-clearance model.

The result is a dense concentration map over the tissue, interpolated from point source locations.
This can be used to:

- Identify where a signalling molecule is most concentrated
- Compute the concentration seen by each cell
- Create two-gene ratio maps (spatial domains of ligand/receptor balance)
- Visualise how diffusion length affects signal range

---

# The physical model

The function solves the **screened Poisson equation** (steady-state diffusion-clearance PDE):

$$D \nabla^2 C(\mathbf{x}) - \lambda C(\mathbf{x}) + s(\mathbf{x}) = 0$$

| Symbol | Meaning | Parameter |
|---|---|---|
| $C(\mathbf{x})$ | Steady-state concentration at position **x** | (what we solve for) |
| $D$ | Diffusion coefficient (spatial spread rate) | `D` |
| $\lambda$ | First-order clearance rate | `lambda` |
| $s(\mathbf{x})$ | Source density (production rate per unit area) | `production_rate` |

The key derived quantity is the **diffusion length**:

$$L = \sqrt{D / \lambda}$$

$L$ sets the characteristic distance over which concentration decays from a point source. It is the
single most important parameter to tune. The shape of the field depends only on $L$; the amplitude
scales as $1/D$.

> **Practical interpretation:** $L$ is roughly the radius of influence of a single transcript. At
> distance $r \gg L$ from a source, the concentration is negligible. At $r \ll L$, concentration is
> near its maximum.

---

# Setup

```{r libraries}
library(TissueField)
library(sf)
library(ggplot2)
library(patchwork)
set.seed(2024)
```

---

# Shared helpers

```{r helpers}
# ---- Plotting theme ----
theme_demo <- function(bs = 9.5)
  theme_minimal(base_size = bs) +
  theme(plot.title    = element_text(face = "bold", size = bs + 0.5),
        plot.subtitle = element_text(size = bs - 1, color = "grey40",
                                     margin = margin(b = 4)),
        panel.grid    = element_line(color = "grey94"),
        legend.key.width  = unit(0.35, "cm"),
        legend.title      = element_text(size = bs - 1),
        legend.text       = element_text(size = bs - 1.5),
        axis.text         = element_text(size = bs - 2))

# ---- Field visualiser ----
plot_field <- function(result, transcripts = NULL,
                       title = "", subtitle = "",
                       palette = "magma", log_scale = FALSE,
                       show_pts = TRUE, show_contours = TRUE, n_contours = 6L,
                       pt_size = 0.35, pt_alpha = 0.45, pt_color = "#00e5ff",
                       fill_label = "Conc.", symmetric = FALSE) {
  df       <- field_to_df(result, inside_only = TRUE)
  fill_col <- if (log_scale) { df$fv <- log1p(df$field); "fv" } else "field"
  fill_lbl <- if (log_scale) paste0("log1p(", fill_label, ")") else fill_label
  p <- ggplot(df, aes(x = x, y = y, fill = .data[[fill_col]])) +
    geom_raster(interpolate = TRUE) + coord_equal() +
    labs(title = title, subtitle = subtitle, x = "X", y = "Y") +
    theme_demo()
  if (symmetric) {
    lim <- max(abs(df[[fill_col]]), na.rm = TRUE)
    p   <- p + scale_fill_distiller(palette = "RdBu", limits = c(-lim, lim),
                                     name = fill_lbl, na.value = "transparent")
  } else {
    p <- p + scale_fill_viridis_c(option = palette, name = fill_lbl,
                                   na.value = "transparent")
  }
  if (show_contours && any(!is.na(df$field)))
    p <- p + geom_contour(aes(z = field), color = "white",
                          alpha = 0.35, linewidth = 0.25, bins = n_contours)
  if (show_pts && !is.null(transcripts))
    p <- p + geom_point(data = transcripts, aes(x = x, y = y),
                        inherit.aes = FALSE, color = pt_color,
                        size = pt_size, alpha = pt_alpha)
  p
}

# ---- Sample points uniformly from inside a mask ----
sample_in_mask <- function(mask_poly, n, max_tries = 20) {
  bb  <- sf::st_bbox(mask_poly); mu <- sf::st_union(mask_poly)
  pts <- data.frame(x = numeric(0), y = numeric(0))
  for (i in seq_len(max_tries)) {
    if (nrow(pts) >= n) break
    cnd <- data.frame(x = runif(n * 8, bb["xmin"], bb["xmax"]),
                      y = runif(n * 8, bb["ymin"], bb["ymax"]))
    ok  <- as.logical(sf::st_within(
      sf::st_as_sf(cnd, coords = c("x","y")), mu, sparse = FALSE)[, 1L])
    pts <- rbind(pts, cnd[ok, ])
  }
  pts[seq_len(min(n, nrow(pts))), ]
}
```

---

# Function reference

```r
estimate_concentration_field(
  mask,               # sfc polygon -- output of fit_spatial_mask()
  transcript_coords,  # data.frame with columns x and y

  # Physics
  D                  = 1.0,      # diffusion coefficient (arbitrary units)
  lambda             = 0.1,      # first-order clearance rate
  diffusion_length   = NULL,     # override: set L directly
  production_rate    = 1.0,

  # Solver
  method             = "fd",     # "fd" | "green" | "kde"
  grid_resolution    = 256L,     # N x N grid
  boundary_condition = "dirichlet",  # "dirichlet" | "neumann"  (fd only)
  fd_solver          = "direct", # "direct" | "iterative"  (fd only)

  # Weighting
  weight_col         = NULL,     # column name for per-transcript weights (e.g. "umi")

  # Optional
  include_external   = FALSE,
  normalize          = FALSE,
  log_transform      = FALSE,
  clip_negative      = TRUE,
  n_cores            = 1L,
  plot               = FALSE,
  verbose            = TRUE
)
```

## Return value

| Element | Contents |
|---|---|
| `field` | N x N matrix; `NA` outside the mask |
| `x`, `y` | Grid centre coordinates (length N each) |
| `hx`, `hy` | Grid cell width and height |
| `mask` | N x N logical matrix (TRUE = inside mask) |
| `sources` | N x N matrix of binned source density |
| `params` | List of all input parameters |
| `diagnostics` | Timing, counts, etc. |

Use `field_to_df(result)` to convert to a tidy data frame for ggplot2.

---

# Section 1 -- Synthetic tissue and transcript layout

A kidney-bean-shaped tissue with two synthetic genes:

- **Gene A**: focal cluster at upper-left, plus sparse background. High UMI count.
- **Gene B**: 500 transcripts with a right-biased spatial distribution. Low UMI count.

```{r sec1-tissue}
in_kidney <- function(x, y, oa = 9, ob = 7, ncx = 5, nr = 3.8)
  (x/oa)^2 + (y/ob)^2 < 1 & sqrt((x - ncx)^2 + y^2) >= nr

cands      <- data.frame(x = runif(18000, -11, 11), y = runif(18000, -9, 9))
tissue_pts <- cands[in_kidney(cands$x, cands$y), ][1:3000, ]

if (requireNamespace("TissueMask", quietly = TRUE)) {
  tissue_mask <- TissueMask::fit_spatial_mask(
    tissue_pts, method = "raster",
    raster_resolution = 256L, raster_sigma = 0.7, raster_threshold = 0.15,
    verbose = FALSE)
} else {
  pts_sf      <- st_as_sf(tissue_pts, coords = c("x","y"))
  tissue_mask <- st_convex_hull(st_union(pts_sf))
}
mu_t <- sf::st_union(tissue_mask)
```

```{r sec1-transcripts}
# Gene A: focal cluster at (-4.5, 3.8) + sparse background
tc_A_cl <- data.frame(x = rnorm(350, -4.5, 1.2), y = rnorm(350, 3.8, 0.9))
A_in    <- as.logical(sf::st_within(
  sf::st_as_sf(tc_A_cl, coords = c("x","y")), mu_t, sparse = FALSE)[, 1L])
tc_A_cl <- tc_A_cl[A_in, ]
tc_A_sc <- sample_in_mask(tissue_mask, 50)
tc_A    <- rbind(tc_A_cl, tc_A_sc)
tc_A$umi <- c(rpois(nrow(tc_A_cl), 15), rpois(nrow(tc_A_sc), 2))

# Gene B: right-biased random distribution
tc_B_all <- sample_in_mask(tissue_mask, 2000)
prob_B   <- plogis(tc_B_all$x / 3)
tc_B     <- tc_B_all[runif(nrow(tc_B_all)) < prob_B, ][1:500, ]
tc_B$umi <- rpois(nrow(tc_B), 5)
```

```{r sec1-plot, fig.height=5}
tissue_sf <- sf::st_sf(geometry = tissue_mask)

ggplot() +
  geom_sf(data = tissue_sf, fill = "grey92", color = "grey60", linewidth = 0.4) +
  geom_point(data = tc_A, aes(x = x, y = y, color = "Gene A", size = log1p(umi)),
             alpha = 0.6) +
  geom_point(data = tc_B, aes(x = x, y = y, color = "Gene B", size = log1p(umi)),
             alpha = 0.5) +
  scale_color_manual(values = c("Gene A" = "#e74c3c", "Gene B" = "#2980b9"),
                     name = NULL) +
  scale_size_continuous(range = c(0.3, 2.5), name = "log(UMI+1)") +
  coord_sf(datum = NA) + theme_demo() +
  labs(title    = "Section 1: Synthetic kidney-bean tissue",
       subtitle = "Gene A: focal upper-left | Gene B: broad right-biased",
       x = "X", y = "Y")
```

---

# Section 2 -- Method comparison: fd, green, kde

```{r sec2}
pb <- list(D = 1, lambda = 0.3, production_rate = 1,
           grid_resolution = 256L, verbose = FALSE)

res_fd    <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A,
         method = "fd", fd_solver = "direct",
         boundary_condition = "dirichlet"), pb))

res_green <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A,
         method = "green"), pb))

res_kde   <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A,
         method = "kde", kde_bandwidth = sqrt(1/0.3)), pb))

res_diff        <- res_fd
res_diff$field  <- res_fd$field - res_green$field
```

```{r sec2-plot, fig.height=8}
(plot_field(res_fd,    tc_A, title = "2a. fd (Dirichlet)",
            subtitle = "Exact BCs; recommended") +
 plot_field(res_green, tc_A, title = "2b. green (FFT)",
            subtitle = "Infinite domain; no boundary effects")) /
(plot_field(res_kde,   tc_A, title = "2c. kde",
            subtitle = "Bandwidth = L; phenomenological") +
 plot_field(res_diff, title = "2d. fd - green",
            subtitle = "fd correctly suppresses near-edge concentration",
            symmetric = TRUE, show_contours = FALSE, show_pts = FALSE)) +
plot_annotation(
  title = "Section 2: Solver Comparison (Gene A, D=1, lambda=0.3, L ~ 1.83)",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**Key observations:**

- `fd` and `green` agree well in the interior; difference map (2d) shows `fd` correctly zeros
  concentration at the tissue boundary (Dirichlet), while `green` leaks outward.
- `kde` is visually similar because the bandwidth was set to $L$, but it has no physical
  interpretation and no boundary conditions.

---

# Section 3 -- Diffusion length sweep

The diffusion length $L = \sqrt{D/\lambda}$ is the primary tuning parameter.

```{r sec3}
L_vals <- c(0.5, 1.5, 4.0, 10.0)

sw3 <- lapply(L_vals, function(Lv) {
  res <- estimate_concentration_field(
    mask = tissue_mask, transcript_coords = tc_A,
    D = 1, diffusion_length = Lv, production_rate = 1,
    method = "fd", fd_solver = "direct", grid_resolution = 256L, verbose = FALSE)
  f         <- res$field
  res$field <- (f - min(f, na.rm = TRUE)) / diff(range(f, na.rm = TRUE))
  plot_field(res, tc_A,
             title      = sprintf("L = %.1f", Lv),
             subtitle   = sprintf("lambda ~ %.3f", 1/Lv^2),
             fill_label = "Norm. Conc.", pt_size = 0.4) +
    scale_fill_viridis_c(option = "magma", name = "Norm.\nConc.",
                          limits = c(0, 1), na.value = "transparent")
})
```

```{r sec3-plot, fig.height=4}
wrap_plots(sw3, nrow = 1) +
  plot_annotation(
    title    = "Section 3: Diffusion Length Sweep (each panel normalised independently)",
    subtitle = "Small L: tight peaks  |  Large L: tissue-wide field",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 9, color = "grey40")))
```

```{r sec3-peaks}
peaks <- sapply(L_vals, function(Lv) {
  res <- estimate_concentration_field(
    mask = tissue_mask, transcript_coords = tc_A,
    D = 1, diffusion_length = Lv,
    method = "fd", fd_solver = "direct", grid_resolution = 256L, verbose = FALSE)
  max(res$field, na.rm = TRUE)
})
knitr::kable(data.frame(L = L_vals, peak_concentration = round(peaks, 5)))
```

### How to choose L in practice

1. **What is the typical inter-cluster distance?** $L$ should be 10--20% of this distance.
2. **Are you modelling a small diffusing molecule or a large one?** Small molecules have higher $D$ and thus larger $L$.
3. **Does your field look biologically reasonable?** Inspect the output and compare to known biology.

---

# Section 4 -- Fixed L, varying D and lambda

**Field shape depends only on $L$.** Changing $D$ and $\lambda$ while keeping $L = \sqrt{D/\lambda}$ constant produces identical spatial patterns, scaling amplitude as $1/D$.

```{r sec4}
DL_cases <- list(
  list(D = 0.1, lam = 0.025, label = "D=0.1, lam=0.025"),
  list(D = 1,   lam = 0.25,  label = "D=1,   lam=0.25"),
  list(D = 5,   lam = 1.25,  label = "D=5,   lam=1.25"),
  list(D = 20,  lam = 5,     label = "D=20,  lam=5")
)

dl_pl <- lapply(DL_cases, function(co) {
  res <- estimate_concentration_field(
    mask = tissue_mask, transcript_coords = tc_A,
    D = co$D, lambda = co$lam, production_rate = 1,
    method = "fd", fd_solver = "direct", grid_resolution = 256L, verbose = FALSE)
  pk <- round(max(res$field, na.rm = TRUE), 5)
  plot_field(res, tc_A,
             title    = co$label,
             subtitle = sprintf("L=2 fixed | peak=%.5f", pk),
             pt_size  = 0.3)
})
```

```{r sec4-plot, fig.height=4}
wrap_plots(dl_pl, nrow = 1) +
  plot_annotation(
    title    = "Section 4: Fixed L=2, Varying D and lambda",
    subtitle = "Shape IDENTICAL (same L). Amplitude proportional to 1/D.",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 9, color = "grey40")))
```

---

# Section 5 -- Boundary conditions

```{r sec5}
res_dir <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.2, method = "fd", fd_solver = "direct",
  boundary_condition = "dirichlet", grid_resolution = 256L, verbose = FALSE)

res_neu <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.2, method = "fd", fd_solver = "direct",
  boundary_condition = "neumann", grid_resolution = 256L, verbose = FALSE)

res_bcd        <- res_dir
res_bcd$field  <- res_neu$field - res_dir$field
```

```{r sec5-crosssection}
sl <- function(res, lbl) {
  df <- field_to_df(res)
  s  <- df[abs(df$y) < res$hy * 1.5 & !is.na(df$field), ]
  s  <- s[order(s$x), ]; s$BC <- lbl; s
}
sl_all <- rbind(sl(res_dir, "Dirichlet"), sl(res_neu, "Neumann"))

p5d <- ggplot(sl_all, aes(x = x, y = field, color = BC)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c("Dirichlet" = "#e74c3c", "Neumann" = "#2980b9")) +
  labs(title    = "5d. Cross-section at y = 0",
       subtitle = "Dirichlet drops to 0; Neumann stays elevated",
       x = "X", y = "Conc.", color = "BC") +
  theme_demo()
```

```{r sec5-plot, fig.height=8}
(plot_field(res_dir, tc_A,
            title    = "5a. Dirichlet (C=0 at boundary)",
            subtitle = "D=1, lambda=0.2, L ~ 2.24") +
 plot_field(res_neu, tc_A,
            title    = "5b. Neumann (dC/dn=0)",
            subtitle = "Molecules reflected; higher near edges",
            palette  = "plasma")) /
(plot_field(res_bcd,
            title     = "5c. Neumann - Dirichlet",
            subtitle  = "Neumann retains more concentration near boundary",
            symmetric = TRUE, show_contours = FALSE, show_pts = FALSE) +
 p5d) +
plot_annotation(
  title = "Section 5: Boundary Condition Comparison",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**Recommendation:** Use `"dirichlet"` for tissue sections. Neumann BCs are mainly useful when comparing to analytic solutions or modelling sealed compartments.

---

# Section 6 -- UMI weighting vs. equal weighting

```{r sec6}
res_uw <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, weight_col = NULL, verbose = FALSE)

res_wt <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, weight_col = "umi", verbose = FALSE)

nrm <- function(r) {
  f <- r$field
  r$field <- (f - min(f, na.rm = TRUE)) / diff(range(f, na.rm = TRUE))
  r
}
```

```{r sec6-plot, fig.height=4}
plot_field(nrm(res_uw), tc_A,
           title      = "6a. Unweighted",
           subtitle   = "All transcripts equal weight",
           fill_label = "Norm. Conc.") +
plot_field(nrm(res_wt), tc_A,
           title      = "6b. UMI-weighted",
           subtitle   = "High-UMI cluster source amplified",
           palette    = "plasma",
           fill_label = "Norm. Conc.") +
plot_annotation(
  title = "Section 6: UMI Weighting",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

---

# Section 7 -- External transcripts

`include_external = TRUE` includes transcripts outside the mask as sources, useful when transcripts
fall just outside the fitted mask boundary.

```{r sec7}
tc_ext <- data.frame(x = rnorm(80, 6.5, 0.4),
                     y = rnorm(80, 0, 0.5),
                     umi = rpois(80, 10))
ext_in  <- as.logical(sf::st_within(
  sf::st_as_sf(tc_ext, coords = c("x","y")), mu_t, sparse = FALSE)[, 1L])
tc_ext  <- tc_ext[!ext_in, ]
tc_Ax   <- rbind(tc_A, tc_ext)

res_ne <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_Ax,
  D = 1, lambda = 0.15, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, include_external = FALSE, verbose = FALSE)

res_we <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_Ax,
  D = 1, lambda = 0.15, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, include_external = TRUE, verbose = FALSE)

ext_pt_layer <- geom_point(data = tc_ext, aes(x = x, y = y),
                            color = "orange", size = 0.9, alpha = 0.8,
                            inherit.aes = FALSE)
```

```{r sec7-plot, fig.height=4}
(plot_field(res_ne, tc_A, title = "7a. include_external = FALSE") +
 ext_pt_layer) +
(plot_field(res_we, tc_A, title = "7b. include_external = TRUE",
            palette = "plasma") +
 ext_pt_layer) +
plot_annotation(
  title = "Section 7: External Transcripts (orange = outside mask)",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

---

# Section 8 -- Holey mask: vascular clearance

A key strength of `method = "fd"` is correct handling of holes (vessel lumens) in the mask. The `green` method ignores topology.

```{r sec8-mask}
vd   <- list(list(cx = -2,  cy =  1.5, r = 1.2),
              list(cx =  2.5, cy = -2,  r = 1.0),
              list(cx = -4.5, cy = -2.5, r = 0.8))
in_v <- function(x, y) Reduce(`|`, lapply(vd, function(v)
  sqrt((x - v$cx)^2 + (y - v$cy)^2) < v$r))

n_h   <- 3000
cnd_h <- data.frame(x = runif(n_h * 8, -11, 11),
                    y = runif(n_h * 8, -9,   9))
kp_h  <- in_kidney(cnd_h$x, cnd_h$y) & !in_v(cnd_h$x, cnd_h$y)
pts_h <- cnd_h[kp_h, ][1:n_h, ]

if (requireNamespace("TissueMask", quietly = TRUE)) {
  holey_mask <- TissueMask::fit_spatial_mask(
    pts_h, method = "raster",
    raster_resolution = 256L, raster_sigma = 0.4,
    raster_threshold  = 0.2, verbose = FALSE)
} else {
  pts_hf     <- st_as_sf(pts_h, coords = c("x","y"))
  holey_mask <- st_convex_hull(st_union(pts_hf))
}
tc_h <- sample_in_mask(holey_mask, 400)
```

```{r sec8-solve}
res_hfd    <- estimate_concentration_field(
  mask = holey_mask, transcript_coords = tc_h,
  D = 1, lambda = 0.15, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, verbose = FALSE)

res_htight <- estimate_concentration_field(
  mask = holey_mask, transcript_coords = tc_h,
  D = 1, lambda = 1.0, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, verbose = FALSE)

res_hgrn   <- estimate_concentration_field(
  mask = holey_mask, transcript_coords = tc_h,
  D = 1, lambda = 0.15, method = "green",
  grid_resolution = 256L, verbose = FALSE)

res_hdiff        <- res_hfd
res_hdiff$field  <- res_hfd$field - res_hgrn$field
```

```{r sec8-plot, fig.height=8}
(plot_field(res_hfd,    tc_h,
            title    = "8a. fd | L ~ 2.58",
            subtitle = "Vessel lumens correctly zeroed",
            pt_size  = 0.3) +
 plot_field(res_htight, tc_h,
            title    = "8b. fd | L = 1 (tight)",
            subtitle = "Steeper depletion near vessels",
            palette  = "inferno", pt_size = 0.3)) /
(plot_field(res_hgrn,  tc_h,
            title    = "8c. green -- topology blind",
            subtitle = "FFT ignores holes entirely",
            palette  = "plasma", pt_size = 0.3) +
 plot_field(res_hdiff,
            title     = "8d. fd - green",
            subtitle  = "fd zeros lumens; green does not",
            symmetric = TRUE, show_contours = FALSE, show_pts = FALSE)) +
plot_annotation(
  title    = "Section 8: Holey Mask -- Vascular Clearance",
  subtitle = "Use method='fd' whenever the mask has holes",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

> **Important:** If your tissue mask has holes (vessel lumens, necrotic cores), always use
> `method = "fd"`. The `green` and `kde` methods are unaware of domain topology.

---

# Section 9 -- Wide dynamic range and log-scale visualisation

When transcript counts span several orders of magnitude, use `log_scale = TRUE` in `plot_field()`
or `log_transform = TRUE` in the solver call to reveal background structure.

```{r sec9}
tc_w <- rbind(
  data.frame(x = rnorm(20, -5, 0.3),  y = rnorm(20, 4, 0.3),  umi = rpois(20, 200)),
  data.frame(x = rnorm(300,  0, 4),   y = rnorm(300, 0, 3),   umi = rpois(300, 1)))
tc_w_in <- as.logical(sf::st_within(
  sf::st_as_sf(tc_w[, c("x","y")], coords = c("x","y")),
  mu_t, sparse = FALSE)[, 1L])
tc_w <- tc_w[tc_w_in, ]

res_w <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_w,
  D = 1, lambda = 0.5, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, weight_col = "umi", verbose = FALSE)
```

```{r sec9-plot, fig.height=4}
plot_field(res_w, tc_w,
           title    = "9a. Linear scale",
           subtitle = "Background invisible; cluster dominates",
           pt_color = "white") +
plot_field(res_w, tc_w,
           log_scale = TRUE,
           title    = "9b. log1p scale",
           subtitle = "Background and cluster both visible",
           palette  = "viridis", pt_color = "white") +
plot_annotation(
  title    = "Section 9: Wide Dynamic Range",
  subtitle = "Use log_scale=TRUE when max/min ratio > ~20",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

---

# Section 10 -- Two-gene ratio map

A `log2(A/B)` ratio map shows spatial domains of ligand/receptor balance.

```{r sec10}
ph <- list(D = 1, lambda = 0.25, production_rate = 1,
           method = "fd", fd_solver = "direct",
           grid_resolution = 256L, verbose = FALSE)

res_gA <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A), ph))
res_gB <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_B), ph))

res_rt        <- res_gA
res_rt$field  <- log2((res_gA$field + 1e-6) / (res_gB$field + 1e-6))
```

```{r sec10-plot, fig.height=8}
p_ratio <- plot_field(res_rt,
  title     = "10c. log2(A/B) ratio",
  subtitle  = "Red=A territory | Blue=B territory | White=balanced",
  symmetric = TRUE, show_contours = TRUE, n_contours = 5,
  show_pts  = FALSE, fill_label = "log2(A/B)") +
  geom_point(data = tc_A, aes(x = x, y = y),
             color = "#e74c3c", size = 0.3, alpha = 0.4,
             inherit.aes = FALSE) +
  geom_point(data = tc_B, aes(x = x, y = y),
             color = "#27ae60", size = 0.3, alpha = 0.4,
             inherit.aes = FALSE)

p_dens <- ggplot(field_to_df(res_rt), aes(x = field,
    fill = ifelse(field < 0, "B dominant", "A dominant"))) +
  geom_density(alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("B dominant" = "#2980b9",
                                "A dominant" = "#c0392b"), name = NULL) +
  labs(title    = "10d. Distribution of log2(A/B)",
       subtitle = "Fraction of tissue in each territory",
       x = "log2(A/B)", y = "Density") +
  theme_demo()

(plot_field(res_gA, tc_A, title = "10a. Gene A field") +
 plot_field(res_gB, tc_B, title = "10b. Gene B field", palette = "viridis")) /
(p_ratio + p_dens) +
plot_annotation(
  title = "Section 10: Two-Gene Ratio Map",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

---

# Section 11 -- 100k transcripts and solver timing

```{r sec11, message=TRUE}
tc_100k <- sample_in_mask(tissue_mask, 100000)
cat(sprintf("Sampled %d transcripts.\n", nrow(tc_100k)))

t1 <- system.time(r1 <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_100k,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, verbose = TRUE))

t2 <- system.time(r2 <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_100k,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "iterative",
  sor_omega = 1.75, grid_resolution = 256L, verbose = TRUE))

t3 <- system.time(r3 <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_100k,
  D = 1, lambda = 0.3, method = "green",
  grid_resolution = 256L, verbose = TRUE))
```

```{r sec11-table}
knitr::kable(data.frame(
  Solver        = c("fd/direct (N=256)", "fd/SOR (N=256)", "green/FFT (N=256)"),
  Time_s        = round(c(t1["elapsed"], t2["elapsed"], t3["elapsed"]), 2),
  Has_BCs       = c("Yes", "Yes (Dirichlet only)", "No"),
  Handles_holes = c("Yes", "Yes", "No"),
  Notes         = c("Sparse LU; recommended up to N=400",
                    "Iterative; use for N > 400",
                    "Fastest; infinite domain only")
))
```

```{r sec11-plot, fig.height=4}
sub <- tc_100k[sample(nrow(tc_100k), 2000), ]
plot_field(r1, sub,
           title    = sprintf("11a. fd/direct (%.1f s)", t1["elapsed"]),
           subtitle = "N=256 | sparse LU",
           pt_size = 0.15, pt_alpha = 0.2) +
plot_field(r2, sub,
           title    = sprintf("11b. fd/SOR (%.1f s)", t2["elapsed"]),
           subtitle = "N=256 | iterative",
           palette  = "plasma", pt_size = 0.15, pt_alpha = 0.2) +
plot_annotation(
  title    = "Section 11: 100k Transcripts -- Solver Comparison",
  subtitle = "2,000 of 100,000 points shown",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

---

# Section 12 -- Quantitative field extraction

```{r sec12}
res_q <- res_fd

interpolate_field <- function(result, query_pts) {
  xg <- result$x; yg <- result$y; f <- result$field
  vapply(seq_len(nrow(query_pts)), function(i) {
    qx <- query_pts$x[i]; qy <- query_pts$y[i]
    xi <- findInterval(qx, xg); yi <- findInterval(qy, yg)
    if (xi < 1 || xi >= length(xg) || yi < 1 || yi >= length(yg))
      return(NA_real_)
    wx <- (qx - xg[xi]) / (xg[xi+1] - xg[xi])
    wy <- (qy - yg[yi]) / (yg[yi+1] - yg[yi])
    (1-wx)*(1-wy)*f[xi,yi]   + wx*(1-wy)*f[xi+1,yi] +
    (1-wx)*wy   *f[xi,yi+1] + wx*wy    *f[xi+1,yi+1]
  }, numeric(1))
}
```

```{r sec12-queries}
queries <- data.frame(
  label = c("Cluster core", "Cluster edge", "Tissue centre",
             "Far right", "Near notch"),
  x = c(-4.5, -3.0, 0.0,  5.0, 3.0),
  y = c( 3.8,  2.5, 0.0,  0.0, 0.0))
queries$concentration <- interpolate_field(res_q, queries)
knitr::kable(queries, digits = 5)
```

```{r sec12-stats}
df_q <- field_to_df(res_q)

cat(sprintf("Mean concentration:\n  Left half  (x < 0): %.5f\n  Right half (x >= 0): %.5f\n",
    mean(df_q$field[df_q$x <  0 & !is.na(df_q$field)]),
    mean(df_q$field[df_q$x >= 0 & !is.na(df_q$field)])))

pk <- which.max(res_q$field)
N  <- length(res_q$x)
cat(sprintf("\nPeak: x=%.2f, y=%.2f, value=%.6f\n",
    res_q$x[((pk-1L) %% N) + 1L],
    res_q$y[((pk-1L) %/% N) + 1L],
    res_q$field[pk]))

total  <- sum(res_q$field[res_q$mask], na.rm = TRUE) * res_q$hx * res_q$hy
theory <- nrow(tc_A) * 1.0 / 0.3
cat(sprintf("\nField integral: %.4f\nTheory (n * r / lambda): %.4f\nAgreement: %.1f%%\n",
    total, theory, 100 * (1 - abs(total - theory) / theory)))
```

```{r sec12-profile, fig.height=3.5}
prof <- df_q[abs(df_q$y) < res_q$hy * 1.5 & !is.na(df_q$field), ]
ggplot(prof[order(prof$x), ], aes(x = x, y = field)) +
  geom_line(color = "#e74c3c", linewidth = 0.8) +
  geom_area(fill = "#e74c3c", alpha = 0.15) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(title    = "12e. Concentration profile along y = 0",
       subtitle = "Cluster peak at x ~ -4.5; near-zero at right edge",
       x = "X", y = "Concentration") +
  theme_demo()
```

---

# Parameter quick-reference

```
estimate_concentration_field() -- parameter guide
--------------------------------------------------------------------

PHYSICS
  diffusion_length  Primary tuning knob. Set directly: L = 0.5 to 10
                    Rule: L ~ 10-20% of inter-cluster distance
  D                 Affects amplitude only (amplitude proportional to 1/D)
  lambda            Affects amplitude only; lambda = D / L^2
  production_rate   Global amplitude scaling; use weight_col for per-cell

SOLVER
  method = "fd"     Default. Use whenever you need accurate BCs or holes.
  method = "green"  Faster for sweeps; ignores domain shape.
  method = "kde"    Quick baseline; no physics.

  fd_solver = "direct"     N <= 400  (sparse LU decomposition)
  fd_solver = "iterative"  N > 400   (red-black SOR)

  boundary_condition = "dirichlet"  C=0 at boundary. Default.
  boundary_condition = "neumann"    dC/dn=0. Sealed system.

  grid_resolution = 256   Explore. 512 for publication quality.

WEIGHTS
  weight_col = "umi"   Use UMI counts as source strength.

POST-PROCESSING
  log_transform = TRUE    Apply log1p() after solving.
  normalize = TRUE        Rescale to [0,1] for comparison.
  clip_negative = TRUE    Remove small numerical negatives. Default on.
--------------------------------------------------------------------
```

---

# Session information

```{r session-info}
sessionInfo()
```
