---
title: "Getting Started with TissueField"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with TissueField}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
```

## What is a concentration field?

Spatial transcriptomics gives us discrete detections: a list of (x, y) coordinates where individual mRNA molecules were found. But many biological processes operate through continuous gradients — ligand concentrations, cytokine fields, metabolite distributions. **TissueField** bridges that gap.

Given a set of transcript coordinates and a tissue mask polygon, `estimate_concentration_field()` computes the **steady-state concentration** that those transcripts would produce as diffusing point sources inside the tissue:

$$D \nabla^2 C(\mathbf{x}) - \lambda C(\mathbf{x}) + s(\mathbf{x}) = 0$$

The output is a dense numeric matrix — a spatial field — defined over the tissue with `NA` outside it.

The single most important parameter is the **diffusion length**:

$$L = \sqrt{D / \lambda}$$

$L$ is roughly the radius of influence of one transcript. Set it to match the biological scale of interest.

---

## A first example

We start by creating a simple circular tissue and placing synthetic transcripts inside it.

```{r tissue}
library(TissueField)
library(sf)
library(ggplot2)

set.seed(2024)

# Circular mask (radius = 10, centre = origin)
theta     <- seq(0, 2 * pi, length.out = 201)
circle_xy <- cbind(10 * cos(theta), 10 * sin(theta))
circle_xy[201, ] <- circle_xy[1, ]   # close exactly (avoid floating point gap)
mask      <- st_sfc(st_polygon(list(circle_xy)))

# 200 transcripts: focal cluster at (-4, 3) and uniform background
tc_cluster <- data.frame(x = rnorm(120, -4, 1.5), y = rnorm(120, 3, 1.2))
tc_bg      <- data.frame(x = runif(80, -9, 9),    y = runif(80, -9, 9))
# keep only those inside the mask
inside_fn  <- function(d) {
  as.logical(st_within(st_as_sf(d, coords = c("x","y")), mask, sparse = FALSE)[,1])
}
tc <- rbind(tc_cluster[inside_fn(tc_cluster), ], tc_bg[inside_fn(tc_bg), ])
```

### Compute the concentration field

Call `estimate_concentration_field()` with a moderate diffusion length. Here we set `L = 3` directly using the `diffusion_length` argument, which is the simplest way to control spatial scale.

```{r solve, message=TRUE}
res <- estimate_concentration_field(
  mask              = mask,
  transcript_coords = tc,
  diffusion_length  = 3,      # L = 3 spatial units
  D                 = 1,      # sets amplitude (shape depends only on L)
  method            = "fd",   # finite-difference, recommended
  grid_resolution   = 128L,
  verbose           = TRUE
)
```

### Inspect the return value

```{r inspect}
names(res)
dim(res$field)        # N x N matrix
res$params$diffusion_length
res$diagnostics
```

### Visualise with `field_to_df()` and ggplot2

`field_to_df()` unpacks the matrix into a tidy data frame:

```{r plot-manual, fig.height=5}
df <- field_to_df(res)  # inside_only = TRUE by default
head(df)

ggplot(df, aes(x = x, y = y, fill = field)) +
  geom_raster(interpolate = TRUE) +
  geom_point(data = tc, aes(x = x, y = y),
             inherit.aes = FALSE, color = "#00e5ff",
             size = 0.6, alpha = 0.6) +
  scale_fill_viridis_c(option = "magma", name = "Conc.") +
  coord_equal() + theme_minimal(base_size = 12) +
  labs(title = "Concentration field (L = 3)",
       subtitle = "Cyan points = transcript locations")
```

### Or use the built-in plot function

`plot_concentration_field()` wraps the same ggplot2 code with sensible defaults:

```{r plot-builtin, fig.height=5}
plot_concentration_field(res, transcript_coords = tc,
                         show_contours = TRUE, n_contours = 6)
```

---

## Understanding diffusion length

The field shape is entirely determined by $L$. Use `sweep_diffusion_length()` to compare several values at once:

```{r sweep, fig.height=5, fig.width=9}
library(patchwork)

L_vals <- c(0.8, 2, 5, 12)
plots  <- lapply(L_vals, function(Lv) {
  res_l <- estimate_concentration_field(
    mask, tc, diffusion_length = Lv, D = 1,
    method = "fd", grid_resolution = 128L, verbose = FALSE
  )
  df_l  <- field_to_df(res_l)
  # normalise each panel independently for visual comparison
  rng   <- range(df_l$field, na.rm = TRUE)
  df_l$field_norm <- (df_l$field - rng[1]) / diff(rng)
  ggplot(df_l, aes(x = x, y = y, fill = field_norm)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(option = "magma", limits = c(0, 1),
                          name = "Norm.\nConc.", na.value = "transparent") +
    coord_equal() + theme_minimal(base_size = 9) +
    labs(title = sprintf("L = %g", Lv), x = NULL, y = NULL)
})

wrap_plots(plots, nrow = 1) +
  plot_annotation(
    title    = "Effect of diffusion length (each panel normalised independently)",
    subtitle = "Small L: tight peaks around sources  |  Large L: tissue-wide field",
    theme    = theme(plot.title    = element_text(face = "bold", size = 12),
                     plot.subtitle = element_text(size = 9, color = "grey40"))
  )
```

**Practical guidance:**

- Set $L$ to **10--20% of the typical inter-cluster distance** to get fields that peak at sources and decay between them.
- Larger $L$ produces smoother, more "territorial" fields where influence extends across the whole tissue.
- $D$ only affects the amplitude ($C \propto 1/D$), not the spatial pattern. See `vignette("parameter-tuning")` for details.

---

## Next steps

- `vignette("solver-guide")`: compare the three solvers (`fd`, `green`, `kde`) and boundary conditions.
- `vignette("parameter-tuning")`: deep dive into $L$, $D$, $\lambda$, UMI weighting, and grid resolution.
- `vignette("complete-reference")`: full 12-section reference covering all features.
