---
title: "Estimating Steady-State Molecular Concentration Fields"
subtitle: "A guide to `estimate_concentration_field()`: physics, solvers, and tuning"
author: "Raredon Laboratory · Yale School of Medicine"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: kate
    code_folding: show
    fig_width: 10
    fig_height: 5
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  cache      = FALSE
)
```

---

# Overview

Spatial transcriptomics data gives us discrete transcript or cell locations, but many biological processes — ligand gradients, cytokine fields, metabolite concentrations — are continuous. `estimate_concentration_field()` bridges this gap by computing the **steady-state concentration field** that a set of point sources (mRNA transcripts, detected proteins) would produce under a physically motivated diffusion-clearance model.

The result is a dense concentration map over the tissue, interpolated from point source locations. This can be used to:

- Identify where a signalling molecule is most concentrated
- Compute the concentration seen by each cell
- Create two-gene ratio maps (spatial domains of ligand/receptor balance)
- Visualise how diffusion length affects signal range

---

# The physical model

The function solves the **screened Poisson equation** (also called the steady-state diffusion-clearance PDE):

$$D \nabla^2 C(\mathbf{x}) - \lambda C(\mathbf{x}) + s(\mathbf{x}) = 0$$

where:

| Symbol | Meaning | Parameter |
|---|---|---|
| $C(\mathbf{x})$ | Steady-state concentration at position **x** | — (what we solve for) |
| $D$ | Diffusion coefficient (spatial spread rate) | `D` |
| $\lambda$ | First-order clearance rate | `lambda` |
| $s(\mathbf{x})$ | Source density (production rate per unit area) | `production_rate` |

The key derived quantity is the **diffusion length**:

$$L = \sqrt{D / \lambda}$$

$L$ sets the characteristic distance over which concentration decays from a point source. It is the single most important parameter to tune. The shape of the field depends only on $L$; the amplitude scales as $1/D$.

> **Practical interpretation:** $L$ is roughly the radius of influence of a single transcript. At distance $r \gg L$ from a source, the concentration is negligible. At $r \ll L$, concentration is near its maximum.

---

# Setup

```{r libraries}
library(sf)
library(ggplot2)
library(patchwork)
source("fit_spatial_mask.R")
source("estimate_concentration_field.R")

set.seed(2024)
```

---

# Shared helpers

```{r helpers}
# ---- Plot theme ----
theme_demo <- function(bs = 9.5)
  theme_minimal(base_size = bs) +
  theme(plot.title    = element_text(face = "bold", size = bs + 0.5),
        plot.subtitle = element_text(size = bs - 1, color = "grey40",
                                     margin = margin(b = 4)),
        panel.grid    = element_line(color = "grey94"),
        legend.key.width  = unit(0.35, "cm"),
        legend.title      = element_text(size = bs - 1),
        legend.text       = element_text(size = bs - 1.5),
        axis.text         = element_text(size = bs - 2))

# ---- Field visualiser ----
# Uses geom_raster (correct for continuous fields on regular grids).
# show_contours overlays iso-concentration lines.
plot_field <- function(result, transcripts = NULL,
                       title = "", subtitle = "",
                       palette = "magma", log_scale = FALSE,
                       show_pts = TRUE, show_contours = TRUE, n_contours = 6L,
                       pt_size = 0.35, pt_alpha = 0.45, pt_color = "#00e5ff",
                       fill_label = "Conc.", symmetric = FALSE) {
  df       <- field_to_df(result, inside_only = TRUE)
  fill_col <- if (log_scale) { df$fv <- log1p(df$field); "fv" } else "field"
  fill_lbl <- if (log_scale) paste0("log\u2081\u208a(", fill_label, ")") else fill_label
  p <- ggplot(df, aes(x = x, y = y, fill = .data[[fill_col]])) +
    geom_raster(interpolate = TRUE) + coord_equal() +
    labs(title = title, subtitle = subtitle, x = "X", y = "Y") +
    theme_demo()
  if (symmetric) {
    lim <- max(abs(df[[fill_col]]), na.rm = TRUE)
    p   <- p + scale_fill_distiller(palette = "RdBu", limits = c(-lim, lim),
                                     name = fill_lbl, na.value = "transparent")
  } else {
    p <- p + scale_fill_viridis_c(option = palette, name = fill_lbl,
                                   na.value = "transparent")
  }
  if (show_contours && any(!is.na(df$field)))
    p <- p + geom_contour(aes(z = field), color = "white",
                          alpha = 0.35, linewidth = 0.25, bins = n_contours)
  if (show_pts && !is.null(transcripts))
    p <- p + geom_point(data = transcripts, aes(x = x, y = y),
                        inherit.aes = FALSE, color = pt_color,
                        size = pt_size, alpha = pt_alpha)
  p
}

# ---- Sample points uniformly from inside a mask ----
sample_in_mask <- function(mask_poly, n, max_tries = 20) {
  bb  <- sf::st_bbox(mask_poly); mu <- sf::st_union(mask_poly)
  pts <- data.frame(x = numeric(0), y = numeric(0))
  for (i in seq_len(max_tries)) {
    if (nrow(pts) >= n) break
    cnd <- data.frame(x = runif(n * 8, bb["xmin"], bb["xmax"]),
                      y = runif(n * 8, bb["ymin"], bb["ymax"]))
    ok  <- as.logical(sf::st_within(
      sf::st_as_sf(cnd, coords = c("x","y")), mu, sparse = FALSE)[, 1L])
    pts <- rbind(pts, cnd[ok, ])
  }
  pts[seq_len(min(n, nrow(pts))), ]
}
```

---

# Function reference: `estimate_concentration_field()`

```r
estimate_concentration_field(
  mask,               # sfc polygon — output of fit_spatial_mask()
  transcript_coords,  # data.frame with columns x and y (plus optional weight_col)

  # Physics
  D                  = 1.0,      # diffusion coefficient (arbitrary units)
  lambda             = 0.1,      # first-order clearance rate
  diffusion_length   = NULL,     # override: set L directly; NULL uses sqrt(D/lambda)
  production_rate    = 1.0,      # multiplied by weights if weight_col is set

  # Solver
  method             = "fd",     # "fd" | "green" | "kde"
  grid_resolution    = 256L,     # grid cells per axis (N × N)
  boundary_condition = "dirichlet",  # "dirichlet" | "neumann"  (fd only)
  fd_solver          = "direct", # "direct" | "iterative"  (fd only)

  # Weighting
  weight_col         = NULL,     # column name for per-transcript weights (e.g. "umi")

  # Optional
  include_external   = FALSE,    # include transcripts outside mask?
  normalize          = FALSE,    # rescale field to [0, 1]?
  log_transform      = FALSE,    # apply log1p() after solving?
  clip_negative      = TRUE,     # set small negatives to 0?
  n_cores            = 1L,
  plot               = FALSE,
  verbose            = TRUE
)
```

## Return value

A named list containing:

| Element | Contents |
|---|---|
| `field` | N×N matrix; `NA` outside the mask |
| `x`, `y` | Grid centre coordinates (length N each) |
| `hx`, `hy` | Grid cell width and height |
| `mask` | N×N logical matrix (TRUE = inside mask) |
| `sources` | N×N matrix of binned source density |
| `params` | List of all input parameters |
| `diagnostics` | Timing, counts, etc. |

Use `field_to_df(result)` to convert to a tidy data frame for ggplot.

---

# Section 1 — Synthetic tissue and transcript layout

We build a kidney-bean-shaped tissue with two synthetic genes:

- **Gene A**: focal cluster at upper-left, plus sparse background. High UMI count (focal source).
- **Gene B**: 500 transcripts with a right-biased spatial distribution. Low UMI count (diffuse source).

```{r sec1-tissue}
in_kidney <- function(x, y, oa = 9, ob = 7, ncx = 5, nr = 3.8)
  (x/oa)^2 + (y/ob)^2 < 1 & sqrt((x - ncx)^2 + y^2) >= nr

cands      <- data.frame(x = runif(18000, -11, 11), y = runif(18000, -9, 9))
tissue_pts <- cands[in_kidney(cands$x, cands$y), ][1:3000, ]

tissue_mask <- fit_spatial_mask(tissue_pts, method = "raster",
  raster_resolution = 256L, raster_sigma = 0.7, raster_threshold = 0.15,
  verbose = FALSE)

mu_t <- sf::st_union(tissue_mask)
```

```{r sec1-transcripts}
# Gene A: focal cluster at (-4.5, 3.8) + sparse background
tc_A_cl <- data.frame(x = rnorm(350, -4.5, 1.2), y = rnorm(350, 3.8, 0.9))
A_in    <- as.logical(sf::st_within(
  sf::st_as_sf(tc_A_cl, coords = c("x","y")), mu_t, sparse = FALSE)[, 1L])
tc_A_cl <- tc_A_cl[A_in, ]
tc_A_sc <- sample_in_mask(tissue_mask, 50)
tc_A    <- rbind(tc_A_cl, tc_A_sc)
tc_A$umi <- c(rpois(nrow(tc_A_cl), 15), rpois(nrow(tc_A_sc), 2))

# Gene B: right-biased random distribution
tc_B_all <- sample_in_mask(tissue_mask, 2000)
prob_B   <- plogis(tc_B_all$x / 3)
tc_B     <- tc_B_all[runif(nrow(tc_B_all)) < prob_B, ][1:500, ]
tc_B$umi <- rpois(nrow(tc_B), 5)
```

```{r sec1-plot, fig.height=5}
tissue_sf <- sf::st_sf(geometry = tissue_mask)

ggplot() +
  geom_sf(data = tissue_sf, fill = "grey92", color = "grey60", linewidth = 0.4) +
  geom_point(data = tc_A, aes(x = x, y = y, color = "Gene A", size = log1p(umi)),
             alpha = 0.6) +
  geom_point(data = tc_B, aes(x = x, y = y, color = "Gene B", size = log1p(umi)),
             alpha = 0.5) +
  scale_color_manual(values = c("Gene A" = "#e74c3c", "Gene B" = "#2980b9"),
                     name = NULL) +
  scale_size_continuous(range = c(0.3, 2.5), name = "log(UMI+1)") +
  coord_sf(datum = NA) + theme_demo() +
  labs(title    = "Section 1: Synthetic kidney-bean tissue",
       subtitle = "Gene A: focal upper-left | Gene B: broad right-biased",
       x = "X", y = "Y")
```

---

# Section 2 — Method comparison: fd, green, kde

Three solvers are available. Each makes different assumptions about the domain:

| Method | How it works | Boundary handling | Best for |
|---|---|---|---|
| `"fd"` | Sparse finite-difference linear system | **Exact** Dirichlet or Neumann BCs | **Default; recommended** |
| `"green"` | FFT convolution with analytic Green's function $K_0(\kappa r)$ | **None** — infinite domain | Fast interior estimates; quick sweeps |
| `"kde"` | Gaussian kernel smoothing | None | Phenomenological baseline |

```{r sec2, cache=TRUE}
pb <- list(D = 1, lambda = 0.3, production_rate = 1,
           grid_resolution = 256L, verbose = FALSE)

res_fd    <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A,
         method = "fd", fd_solver = "direct",
         boundary_condition = "dirichlet"), pb))

res_green <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A,
         method = "green"), pb))

res_kde   <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A,
         method = "kde", kde_bandwidth = sqrt(1/0.3)), pb))

res_diff  <- res_fd
res_diff$field <- res_fd$field - res_green$field
```

```{r sec2-plot, fig.height=8}
(plot_field(res_fd,    tc_A, title = "2a. fd (Dirichlet)",
            subtitle = "Exact BCs; recommended") +
 plot_field(res_green, tc_A, title = "2b. green (FFT)",
            subtitle = "Infinite domain; no boundary effects")) /
(plot_field(res_kde,   tc_A, title = "2c. kde",
            subtitle = "Bandwidth = L; phenomenological") +
 plot_field(res_diff, title = "2d. fd − green",
            subtitle = "fd correctly suppresses near-edge concentration",
            symmetric = TRUE, show_contours = FALSE, show_pts = FALSE)) +
plot_annotation(
  title = "Section 2: Solver Comparison (Gene A, D=1, λ=0.3, L≈1.83)",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**Key observations:**

- The `fd` and `green` maps look very similar in the interior (panel 2d shows only small differences away from the edges). This confirms that the physics is consistent between methods.
- The **difference map** (2d) shows that `fd` correctly drives concentration to zero at the tissue boundary (Dirichlet BC), while `green` ignores the boundary entirely. This matters most when sources are near the edge.
- `kde` produces a visually similar pattern to `fd` because the bandwidth was set equal to $L$, but it has no physical interpretation and no BCs.

**When to use `method = "green"`:** When you are doing a rapid parameter sweep over many $L$ values and don't need exact boundary conditions. It runs faster than `fd` at high resolution.

---

# Section 3 — Diffusion length sweep: the most important parameter

The diffusion length $L = \sqrt{D/\lambda}$ is the primary parameter to tune. It determines how far each transcript's signal spreads. Small $L$: fields are tight and peaked around clusters. Large $L$: fields fill the entire tissue.

```{r sec3, cache=TRUE}
L_vals <- c(0.5, 1.5, 4.0, 10.0)

sw3 <- lapply(L_vals, function(Lv) {
  res <- estimate_concentration_field(
    mask = tissue_mask, transcript_coords = tc_A,
    D = 1, diffusion_length = Lv, production_rate = 1,
    method = "fd", fd_solver = "direct", grid_resolution = 256L,
    verbose = FALSE)
  # Normalise each panel independently for visual comparison
  f <- res$field
  res$field <- (f - min(f, na.rm = TRUE)) / diff(range(f, na.rm = TRUE))
  plot_field(res, tc_A,
             title      = sprintf("L = %.1f", Lv),
             subtitle   = sprintf("λ ≈ %.3f", 1/Lv^2),
             fill_label = "Norm. Conc.", pt_size = 0.4) +
  scale_fill_viridis_c(option = "magma", name = "Norm.\nConc.",
                        limits = c(0, 1), na.value = "transparent")
})
```

```{r sec3-plot, fig.height=4}
wrap_plots(sw3, nrow = 1) +
  plot_annotation(
    title    = "Section 3: Diffusion Length Sweep (each panel normalised independently)",
    subtitle = "Small L → tight peaks | Large L → tissue-wide field",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 9, color = "grey40")))
```

```{r sec3-peaks}
peaks <- sapply(L_vals, function(Lv) {
  res <- estimate_concentration_field(
    mask = tissue_mask, transcript_coords = tc_A,
    D = 1, diffusion_length = Lv,
    method = "fd", fd_solver = "direct", grid_resolution = 256L,
    verbose = FALSE)
  max(res$field, na.rm = TRUE)
})
data.frame(L = L_vals, peak_concentration = round(peaks, 5))
```

### How to choose L in practice

There is no single correct answer — the right $L$ is biologically motivated. Ask:

1. **What is the typical inter-cluster distance?** $L$ should be 10–20% of this distance if you want fields that peak at sources and drop between them. Larger values produce smoother, more "territorial" fields.
2. **Are you modelling a small diffusing molecule or a large one?** Small molecules (e.g. metabolites, small cytokines) have higher $D$ and thus larger $L$. Large molecules (e.g. ECM-bound ligands) have small $D$.
3. **Does your field look biologically reasonable?** Inspect the output and ask whether the gradients match known biology.

> **Shortcut:** set `diffusion_length = L` directly to bypass the `D`/`lambda` decomposition. The shape depends only on $L$; $D$ only affects amplitude.

---

# Section 4 — Fixed L, varying D and lambda

This section makes the key mathematical point: **field shape is determined entirely by $L$**. Changing $D$ and $\lambda$ while keeping $L = \sqrt{D/\lambda}$ constant produces identical spatial patterns, only scaling the amplitude by $1/D$.

```{r sec4, cache=TRUE}
DL_cases <- list(
  list(D = 0.1, lam = 0.025, label = "D=0.1, λ=0.025"),
  list(D = 1,   lam = 0.25,  label = "D=1,   λ=0.25"),
  list(D = 5,   lam = 1.25,  label = "D=5,   λ=1.25"),
  list(D = 20,  lam = 5,     label = "D=20,  λ=5")
)

dl_pl <- lapply(DL_cases, function(co) {
  res <- estimate_concentration_field(
    mask = tissue_mask, transcript_coords = tc_A,
    D = co$D, lambda = co$lam, production_rate = 1,
    method = "fd", fd_solver = "direct", grid_resolution = 256L,
    verbose = FALSE)
  pk <- round(max(res$field, na.rm = TRUE), 5)
  plot_field(res, tc_A,
             title    = co$label,
             subtitle = sprintf("L=2 fixed | peak=%.5f", pk),
             pt_size  = 0.3)
})
```

```{r sec4-plot, fig.height=4}
wrap_plots(dl_pl, nrow = 1) +
  plot_annotation(
    title    = "Section 4: Fixed L=2, Varying D and λ",
    subtitle = "Shape IDENTICAL (same L). Amplitude ∝ 1/D.",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 9, color = "grey40")))
```

**Implication for parameter setting:** When fitting to data, first determine the spatial extent of influence ($L$), then fix either $D$ or $\lambda$ based on biophysical priors. The third parameter is then determined.

---

# Section 5 — Boundary conditions

The `fd` method supports two boundary conditions:

- **Dirichlet** (`boundary_condition = "dirichlet"`): $C = 0$ at the tissue boundary. Physically: molecules are absorbed or degrade instantly at the boundary. This is appropriate for most spatial transcriptomics sections where the tissue edge is a hard barrier.
- **Neumann** (`boundary_condition = "neumann"`): $\partial C / \partial n = 0$ at the boundary (zero normal flux). Physically: no molecules cross the boundary — they are reflected back in. This is appropriate when modelling a sealed container.

```{r sec5, cache=TRUE}
res_dir <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.2, method = "fd", fd_solver = "direct",
  boundary_condition = "dirichlet", grid_resolution = 256L, verbose = FALSE)

res_neu <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.2, method = "fd", fd_solver = "direct",
  boundary_condition = "neumann", grid_resolution = 256L, verbose = FALSE)

res_bcd <- res_dir
res_bcd$field <- res_neu$field - res_dir$field
```

```{r sec5-crosssection}
sl <- function(res, lbl) {
  df <- field_to_df(res)
  s  <- df[abs(df$y) < res$hy * 1.5 & !is.na(df$field), ]
  s  <- s[order(s$x), ]; s$BC <- lbl; s
}
sl_all <- rbind(sl(res_dir, "Dirichlet"), sl(res_neu, "Neumann"))

p5d <- ggplot(sl_all, aes(x = x, y = field, color = BC)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c("Dirichlet" = "#e74c3c",
                                 "Neumann"   = "#2980b9")) +
  labs(title    = "5d. Cross-section at y ≈ 0",
       subtitle = "Dirichlet drops to 0; Neumann stays elevated",
       x = "X", y = "Conc.", color = "BC") +
  theme_demo()
```

```{r sec5-plot, fig.height=8}
(plot_field(res_dir, tc_A,
            title    = "5a. Dirichlet (C=0 at boundary)",
            subtitle = "D=1, λ=0.2, L≈2.24") +
 plot_field(res_neu, tc_A,
            title    = "5b. Neumann (∂C/∂n=0)",
            subtitle = "Molecules reflected; higher near edges",
            palette  = "plasma")) /
(plot_field(res_bcd,
            title    = "5c. Neumann − Dirichlet",
            subtitle = "Neumann retains more concentration near boundary",
            symmetric = TRUE, show_contours = FALSE, show_pts = FALSE) +
 p5d) +
plot_annotation(
  title = "Section 5: Boundary Condition Comparison",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**Recommendation:** Use `"dirichlet"` for tissue sections. Neumann BCs are mainly useful when comparing to analytic solutions or modelling sealed compartments.

---

# Section 6 — UMI weighting vs. equal weighting

Each transcript can carry a weight (e.g. its UMI count) via the `weight_col` argument. This means a cell with 50 detected transcripts of a gene contributes 50× more source density than a cell with 1.

```{r sec6, cache=TRUE}
res_uw <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, weight_col = NULL, verbose = FALSE)

res_wt <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_A,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, weight_col = "umi", verbose = FALSE)

# Normalise both for visual comparison
nrm <- function(r) {
  f <- r$field
  r$field <- (f - min(f, na.rm=TRUE)) / diff(range(f, na.rm=TRUE))
  r
}
```

```{r sec6-plot, fig.height=4}
plot_field(nrm(res_uw), tc_A,
           title      = "6a. Unweighted",
           subtitle   = "All transcripts equal weight",
           fill_label = "Norm. Conc.") +
plot_field(nrm(res_wt), tc_A,
           title      = "6b. UMI-weighted",
           subtitle   = "High-UMI cluster source amplified",
           palette    = "plasma",
           fill_label = "Norm. Conc.") +
plot_annotation(
  title = "Section 6: UMI Weighting",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**When to use UMI weighting:** When transcript counts meaningfully reflect expression level differences. Note that this amplifies both true signal and noise — noisy high-UMI outliers can dominate the field. Inspect your UMI distribution before applying weights.

---

# Section 7 — External transcripts

`include_external = TRUE` bins transcripts that fall outside the mask into the nearest inside-edge grid cell, allowing them to contribute to the field. This is useful when you have transcripts just outside the mask boundary (e.g. due to mask fitting inaccuracy) that you still want to include.

```{r sec7, cache=TRUE}
tc_ext <- data.frame(x = rnorm(80, 6.5, 0.4),
                      y = rnorm(80, 0, 0.5),
                      umi = rpois(80, 10))
ext_in <- as.logical(sf::st_within(
  sf::st_as_sf(tc_ext, coords = c("x","y")), mu_t, sparse = FALSE)[, 1L])
tc_ext  <- tc_ext[!ext_in, ]   # keep only those outside the mask
tc_Ax   <- rbind(tc_A, tc_ext)

res_ne <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_Ax,
  D = 1, lambda = 0.15, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, include_external = FALSE, verbose = FALSE)

res_we <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_Ax,
  D = 1, lambda = 0.15, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, include_external = TRUE, verbose = FALSE)

ext_pt_layer <- geom_point(data = tc_ext, aes(x = x, y = y),
                             color = "orange", size = 0.9, alpha = 0.8,
                             inherit.aes = FALSE)
```

```{r sec7-plot, fig.height=4}
(plot_field(res_ne, tc_A, title = "7a. include_external = FALSE") +
 ext_pt_layer) +
(plot_field(res_we, tc_A, title = "7b. include_external = TRUE",
            palette = "plasma") +
 ext_pt_layer) +
plot_annotation(
  title = "Section 7: External Transcripts (orange = outside mask)",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

---

# Section 8 — Holey mask: vascular clearance

This section demonstrates a key strength of `method = "fd"`: it correctly handles holes (e.g. vessel lumens) within the tissue mask. The `fd` solver zeroes the concentration inside each hole, modelling vessels as clearance sinks. The `green` FFT method ignores topology entirely — it treats the holes as normal tissue.

```{r sec8-mask, cache=TRUE}
vd    <- list(list(cx = -2,  cy =  1.5, r = 1.2),
               list(cx =  2.5, cy = -2,   r = 1.0),
               list(cx = -4.5, cy = -2.5, r = 0.8))
in_v  <- function(x, y) Reduce(`|`, lapply(vd, function(v)
  sqrt((x - v$cx)^2 + (y - v$cy)^2) < v$r))

n_h   <- 3000
cnd_h <- data.frame(x = runif(n_h * 8, -11, 11),
                     y = runif(n_h * 8, -9,   9))
kp_h  <- in_kidney(cnd_h$x, cnd_h$y) & !in_v(cnd_h$x, cnd_h$y)
pts_h <- cnd_h[kp_h, ][1:n_h, ]

holey_mask <- fit_spatial_mask(pts_h, method = "raster",
  raster_resolution = 256L, raster_sigma = 0.4,
  raster_threshold  = 0.2,  verbose = FALSE)

tc_h <- sample_in_mask(holey_mask, 400)
```

```{r sec8-solve, cache=TRUE}
res_hfd    <- estimate_concentration_field(
  mask = holey_mask, transcript_coords = tc_h,
  D = 1, lambda = 0.15, method = "fd", fd_solver = "direct",
  grid_resolution = 300L, verbose = FALSE)

res_htight <- estimate_concentration_field(
  mask = holey_mask, transcript_coords = tc_h,
  D = 1, lambda = 1.0, method = "fd", fd_solver = "direct",
  grid_resolution = 300L, verbose = FALSE)

res_hgrn   <- estimate_concentration_field(
  mask = holey_mask, transcript_coords = tc_h,
  D = 1, lambda = 0.15, method = "green",
  grid_resolution = 300L, verbose = FALSE)

res_hdiff  <- res_hfd
res_hdiff$field <- res_hfd$field - res_hgrn$field
```

```{r sec8-plot, fig.height=8}
(plot_field(res_hfd,    tc_h,
            title    = "8a. fd | L ≈ 2.58",
            subtitle = "Vessel lumens correctly zeroed",
            pt_size  = 0.3) +
 plot_field(res_htight, tc_h,
            title    = "8b. fd | L = 1 (tight)",
            subtitle = "Steeper depletion near vessels",
            palette  = "inferno", pt_size = 0.3)) /
(plot_field(res_hgrn,  tc_h,
            title    = "8c. green — topology blind",
            subtitle = "FFT ignores holes entirely",
            palette  = "plasma", pt_size = 0.3) +
 plot_field(res_hdiff,
            title    = "8d. fd − green",
            subtitle = "fd zeros lumens; green does not",
            symmetric = TRUE, show_contours = FALSE,
            show_pts  = FALSE)) +
plot_annotation(
  title    = "Section 8: Holey Mask — Vascular Clearance",
  subtitle = "Use method='fd' whenever the mask has holes",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

> **Important:** If your tissue mask has holes (vessel lumens, necrotic cores), always use `method = "fd"`. The `green` and `kde` methods are unaware of domain topology.

---

# Section 9 — Wide dynamic range and log-scale visualisation

When transcript counts span several orders of magnitude (e.g. a dense focal cluster alongside sparse background), the linear concentration scale will be dominated by the peak and the background will appear flat. The `log1p` transform reveals structure at both ends of the dynamic range.

```{r sec9, cache=TRUE}
tc_w <- rbind(
  data.frame(x = rnorm(20, -5, 0.3),  y = rnorm(20, 4, 0.3),  umi = rpois(20, 200)),
  data.frame(x = rnorm(300,  0, 4),   y = rnorm(300, 0, 3),   umi = rpois(300, 1)))
tc_w_in <- as.logical(sf::st_within(
  sf::st_as_sf(tc_w[, c("x","y")], coords = c("x","y")),
  mu_t, sparse = FALSE)[, 1L])
tc_w <- tc_w[tc_w_in, ]

res_w <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_w,
  D = 1, lambda = 0.5, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, weight_col = "umi", verbose = FALSE)
```

```{r sec9-plot, fig.height=4}
plot_field(res_w, tc_w,
           title    = "9a. Linear scale",
           subtitle = "Background invisible; cluster dominates",
           pt_color = "white") +
plot_field(res_w, tc_w,
           log_scale = TRUE,
           title    = "9b. log₁₊ scale",
           subtitle = "Background and cluster both visible",
           palette  = "viridis", pt_color = "white") +
plot_annotation(
  title    = "Section 9: Wide Dynamic Range",
  subtitle = "Use log_scale=TRUE when max/min UMI ratio > ~20",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

You can also apply the log transform inside the solver with `log_transform = TRUE`, which is equivalent but applies the transform before returning the field matrix.

---

# Section 10 — Two-gene ratio map

A **log₂(A/B) ratio map** shows where Gene A has higher concentration relative to Gene B, and vice versa. This is analogous to a log-fold-change map, but in space rather than between conditions. Blue regions favour B; yellow/red regions favour A; white is balanced.

```{r sec10, cache=TRUE}
ph <- list(D = 1, lambda = 0.25, production_rate = 1,
           method = "fd", fd_solver = "direct",
           grid_resolution = 256L, verbose = FALSE)

res_gA <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_A), ph))
res_gB <- do.call(estimate_concentration_field,
  c(list(mask = tissue_mask, transcript_coords = tc_B), ph))

res_rt        <- res_gA
res_rt$field  <- log2((res_gA$field + 1e-6) / (res_gB$field + 1e-6))
```

```{r sec10-plot, fig.height=8}
p_ratio <- plot_field(res_rt,
  title    = "10c. log₂(A/B) ratio",
  subtitle = "Red=A territory | Blue=B territory | White=balanced",
  symmetric = TRUE, show_contours = TRUE, n_contours = 5,
  show_pts = FALSE, fill_label = "log₂(A/B)") +
  geom_point(data = tc_A, aes(x = x, y = y),
             color = "#e74c3c", size = 0.3, alpha = 0.4,
             inherit.aes = FALSE) +
  geom_point(data = tc_B, aes(x = x, y = y),
             color = "#27ae60", size = 0.3, alpha = 0.4,
             inherit.aes = FALSE)

p_dens <- ggplot(field_to_df(res_rt), aes(x = field,
    fill = ifelse(field < 0, "B dominant", "A dominant"))) +
  geom_density(alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("B dominant" = "#2980b9",
                                "A dominant" = "#c0392b"),
                    name = NULL) +
  labs(title    = "10d. Distribution of log₂(A/B)",
       subtitle = "Fraction of tissue in each territory",
       x = "log₂(A/B)", y = "Density") +
  theme_demo()

(plot_field(res_gA, tc_A, title = "10a. Gene A field") +
 plot_field(res_gB, tc_B, title = "10b. Gene B field", palette = "viridis")) /
(p_ratio + p_dens) +
plot_annotation(
  title = "Section 10: Two-Gene Ratio Map",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

The `1e-6` pseudocount in the ratio calculation prevents division-by-zero and log(0) in regions where one gene has near-zero concentration.

---

# Section 11 — 100k transcripts and solver timing

This section benchmarks the three solvers on 100,000 transcripts at two grid resolutions.

```{r sec11, cache=TRUE}
tc_100k <- sample_in_mask(tissue_mask, 100000)
cat(sprintf("Sampled %d transcripts.\n", nrow(tc_100k)))

t1 <- system.time(r1 <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_100k,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "direct",
  grid_resolution = 256L, verbose = TRUE))

t2 <- system.time(r2 <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_100k,
  D = 1, lambda = 0.3, method = "fd", fd_solver = "iterative",
  sor_omega = 1.75, grid_resolution = 512L, verbose = TRUE))

t3 <- system.time(r3 <- estimate_concentration_field(
  mask = tissue_mask, transcript_coords = tc_100k,
  D = 1, lambda = 0.3, method = "green",
  grid_resolution = 512L, verbose = TRUE))
```

```{r sec11-table}
data.frame(
  Solver          = c("fd/direct (N=256)", "fd/SOR (N=512)", "green/FFT (N=512)"),
  Time_s          = round(c(t1["elapsed"], t2["elapsed"], t3["elapsed"]), 2),
  Has_BCs         = c("Yes", "Yes", "No"),
  Handles_holes   = c("Yes", "Yes", "No"),
  Notes           = c("Sparse LU; recommended up to N=400",
                       "Iterative; use for N > 400",
                       "Fastest; infinite domain only")
)
```

```{r sec11-plot, fig.height=4}
sub <- tc_100k[sample(nrow(tc_100k), 2000), ]
plot_field(r1, sub,
           title    = sprintf("11a. fd/direct (%.1f s)", t1["elapsed"]),
           subtitle = "N=256 | sparse LU",
           pt_size = 0.15, pt_alpha = 0.2) +
plot_field(r2, sub,
           title    = sprintf("11b. fd/SOR (%.1f s)", t2["elapsed"]),
           subtitle = "N=512 | iterative",
           palette  = "plasma", pt_size = 0.15, pt_alpha = 0.2) +
plot_field(r3, sub,
           title    = sprintf("11c. green/FFT (%.1f s)", t3["elapsed"]),
           subtitle = "N=512 | FFT",
           palette  = "viridis", pt_size = 0.15, pt_alpha = 0.2) +
plot_annotation(
  title    = "Section 11: 100k Transcripts — Solver Comparison",
  subtitle = "2,000 of 100,000 points shown",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

### Solver selection guide

| Situation | Recommended solver |
|---|---|
| N ≤ 400, holey mask, or exact BCs needed | `fd`, `fd_solver = "direct"` |
| N > 400, high-res output needed | `fd`, `fd_solver = "iterative"` (SOR) |
| Fast sweep over L values, no BCs | `"green"` |
| Quick visual check | `"kde"` |

---

# Section 12 — Quantitative field extraction

Beyond visualisation, the field output can be queried numerically. Here we demonstrate the main extraction operations.

```{r sec12}
res_q <- res_fd   # reuse the Gene A result from Section 2

# ---- Bilinear interpolation at arbitrary query points ----
interpolate_field <- function(result, query_pts) {
  xg <- result$x; yg <- result$y; f <- result$field
  vapply(seq_len(nrow(query_pts)), function(i) {
    qx <- query_pts$x[i]; qy <- query_pts$y[i]
    xi <- findInterval(qx, xg); yi <- findInterval(qy, yg)
    if (xi < 1 || xi >= length(xg) || yi < 1 || yi >= length(yg))
      return(NA_real_)
    wx <- (qx - xg[xi]) / (xg[xi+1] - xg[xi])
    wy <- (qy - yg[yi]) / (yg[yi+1] - yg[yi])
    (1-wx)*(1-wy)*f[xi,yi]   + wx*(1-wy)*f[xi+1,yi] +
    (1-wx)*wy   *f[xi,yi+1] + wx*wy    *f[xi+1,yi+1]
  }, numeric(1))
}
```

```{r sec12-queries}
queries <- data.frame(
  label = c("Cluster core", "Cluster edge", "Tissue centre",
             "Far right", "Near notch"),
  x = c(-4.5, -3.0, 0.0,  5.0, 3.0),
  y = c( 3.8,  2.5, 0.0,  0.0, 0.0))
queries$concentration <- interpolate_field(res_q, queries)
queries
```

```{r sec12-stats}
df_q <- field_to_df(res_q)

# 12b: Mean concentration by tissue half
cat(sprintf("Mean concentration:\n  Left half  (x < 0): %.5f\n  Right half (x ≥ 0): %.5f\n",
    mean(df_q$field[df_q$x <  0 & !is.na(df_q$field)]),
    mean(df_q$field[df_q$x >= 0 & !is.na(df_q$field)])))

# 12c: Peak location
pk <- which.max(res_q$field)
N  <- length(res_q$x)
cat(sprintf("\nPeak: x=%.2f, y=%.2f, value=%.6f\n",
    res_q$x[((pk-1L) %% N) + 1L],
    res_q$y[((pk-1L) %/% N) + 1L],
    res_q$field[pk]))

# 12d: Field integral
total  <- sum(res_q$field[res_q$mask], na.rm = TRUE) * res_q$hx * res_q$hy
theory <- nrow(tc_A) * 1.0 / 0.3   # n * production_rate / lambda
cat(sprintf("\nField integral: %.4f\nTheory (n·r/λ): %.4f\nAgreement: %.1f%%\n",
    total, theory, 100 * (1 - abs(total - theory) / theory)))
```

```{r sec12-profile, fig.height=3.5}
prof <- df_q[abs(df_q$y) < res_q$hy * 1.5 & !is.na(df_q$field), ]
ggplot(prof[order(prof$x), ], aes(x = x, y = field)) +
  geom_line(color = "#e74c3c", linewidth = 0.8) +
  geom_area(fill = "#e74c3c", alpha = 0.15) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(title    = "12e. Concentration profile along y ≈ 0",
       subtitle = "Cluster peak at x ≈ −4.5; near-zero at right edge",
       x = "X", y = "Concentration") +
  theme_demo()
```

---

# Parameter quick-reference

```
estimate_concentration_field() — parameter guide
─────────────────────────────────────────────────────────────────────

PHYSICS
  diffusion_length  Primary tuning knob. Set directly: L = 0.5 to 10
                    Rule: L ~ 10–20% of inter-cluster distance
  D                 Affects amplitude only (amplitude ∝ 1/D)
  lambda            Affects amplitude only; lambda = D / L^2
  production_rate   Global amplitude scaling; use weight_col for per-cell

SOLVER
  method = "fd"     Default. Use whenever you need accurate BCs or holes.
  method = "green"  Faster for sweeps; ignores domain shape.
  method = "kde"    Quick baseline; no physics.

  fd_solver = "direct"     N ≤ 400  (sparse LU decomposition)
  fd_solver = "iterative"  N > 400  (red-black SOR)

  boundary_condition = "dirichlet"  C=0 at boundary. Default. Tissue sections.
  boundary_condition = "neumann"    dC/dn=0. Sealed system.

  grid_resolution = 256   Explore. 512 for publication quality.

WEIGHTS
  weight_col = "umi"   Use UMI counts as source strength.
                        Careful: amplifies outliers.

POST-PROCESSING
  log_transform = TRUE    Apply log1p() after solving. Useful for wide dynamic range.
  normalize = TRUE        Rescale to [0,1] for comparison across conditions.
  clip_negative = TRUE    Remove small numerical negatives. Default on.
─────────────────────────────────────────────────────────────────────
```

---

# Session information

```{r session-info}
sessionInfo()
```
