---
title: "Fitting Spatial Masks to Point Coordinate Data"
subtitle: "A guide to `fit_spatial_mask()`: methods, parameters, and tuning"
author: "Raredon Laboratory · Yale School of Medicine"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: kate
    code_folding: show
    fig_width: 10
    fig_height: 5
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  cache      = FALSE
)
```

---

# Overview

In spatial transcriptomics and spatial proteomics experiments, the tissue of interest occupies some irregular region of a slide. Before computing any spatially-resolved quantity (cell-type composition, signaling gradients, niche structure), you need a reliable **spatial mask**: a polygon or multi-polygon that faithfully represents the shape of the tissue, including internal voids (vessel lumens, necrotic cores, tissue tears) and disconnected fragments.

`fit_spatial_mask()` takes a set of XY point coordinates — typically cell centroids or transcript locations — and fits a mask polygon to them. It supports four methods of increasing complexity:

| Method | Topology | Speed | Best for |
|---|---|---|---|
| `"convex"` | No holes, no islands | Instant | Quick sanity checks |
| `"concave"` | No holes, no islands | Fast | Simple non-convex shapes |
| `"kde"` | Holes and islands | Moderate | Dense, smooth datasets |
| `"raster"` | Holes and islands | Fast at scale | Most use cases; recommended default |

---

# Setup

Place `fit_spatial_mask.R` in your working directory, then source it. The packages below are required.

```{r libraries}
library(sf)
library(ggplot2)
library(patchwork)
source("fit_spatial_mask.R")

set.seed(42)
```

---

# Shared plotting helper

Throughout this vignette we use a single `plot_mask()` helper. The key implementation note is that it uses `geom_sf()` rather than `geom_polygon()`. This is essential: `geom_polygon()` draws a line between the last vertex of the exterior ring and the first vertex of each interior (hole) ring, producing diagonal artefacts. `geom_sf()` delegates rendering to the sf/GEOS layer which handles holes correctly.

```{r plot-helper}
plot_mask <- function(mask, coords, title = "", subtitle = "",
                      fill = "#4c9be8", border = "#1a5fa8",
                      pt_col = "#c0392b", pt_size = 1.2, pt_alpha = 0.55) {
  mask_sf <- sf::st_sf(geometry = mask)
  ggplot() +
    geom_sf(data = mask_sf, fill = fill, alpha = 0.2,
            color = border, linewidth = 0.55) +
    geom_point(data = coords, aes(x = x, y = y),
               color = pt_col, size = pt_size, alpha = pt_alpha) +
    coord_sf(datum = NA) +   # suppress geographic graticule
    labs(title = title, subtitle = subtitle, x = "X", y = "Y") +
    theme_minimal(base_size = 9.5) +
    theme(plot.title    = element_text(face = "bold", size = 9.5),
          plot.subtitle = element_text(size = 8, color = "grey40"),
          panel.grid    = element_line(color = "grey93"))
}
```

---

# Function reference: `fit_spatial_mask()`

```r
fit_spatial_mask(
  coords,

  # Method selection
  method            = "raster",   # "convex" | "concave" | "kde" | "raster"

  # Concave hull
  concavity         = 2,          # lower = tighter; higher = approaches convex
  length_threshold  = 0,

  # KDE
  kde_bandwidth     = NULL,       # NULL = auto (bandwidth.nrd)
  kde_threshold     = 0.05,       # quantile of density to threshold at
  kde_resolution    = 256L,

  # Raster (recommended)
  raster_resolution = 256L,       # grid cells per axis
  raster_sigma      = NULL,       # Gaussian spread in COORDINATE UNITS; NULL = 3% of domain
  raster_threshold  = 0.15,       # fraction of peak Gaussian field to threshold at
  raster_min_pts    = 1L,         # min points per cell to count as occupied

  # Post-processing
  buffer_dist       = 0,          # expand mask outward by this distance
  smooth_mask       = FALSE,      # simplify boundary (st_simplify)
  smooth_tolerance  = NULL,       # NULL = 1% of bounding box diagonal

  # Misc
  n_cores           = 1L,
  crs               = NA,
  plot              = FALSE,
  verbose           = TRUE
)
```

## Return value

`fit_spatial_mask()` returns an `sfc` geometry object (the `sf` package's geometry column class). You can:

- Pass it directly to `estimate_concentration_field()` as the `mask` argument
- Plot it with `geom_sf()`
- Compute area with `sf::st_area()`
- Test point containment with `sf::st_within()`
- Convert to a data frame with `as.data.frame(sf::st_coordinates(mask))`

---

# Section 1 — Circular cloud: sanity check

The simplest possible test: a Gaussian cloud with no special topology. All three non-trivial methods should return similar results here. This section establishes that the function runs and that the three main methods agree on simple inputs.

```{r sec1-data}
n              <- 400
coords_circle  <- data.frame(x = rnorm(n, 0, 1),
                              y = rnorm(n, 0, 1))
```

```{r sec1-fit}
m1a <- fit_spatial_mask(coords_circle, method = "convex",  verbose = FALSE)
m1b <- fit_spatial_mask(coords_circle, method = "concave", concavity = 2,
                         verbose = FALSE)
m1c <- fit_spatial_mask(coords_circle, method = "kde",
                         kde_threshold = 0.02, buffer_dist = 0.1,
                         verbose = FALSE)
```

```{r sec1-plot, fig.height=4}
plot_mask(m1a, coords_circle, title = "1a. Convex hull",
          subtitle = "method='convex'") +
plot_mask(m1b, coords_circle, title = "1b. Concave hull",
          subtitle = "concavity=2") +
plot_mask(m1c, coords_circle, title = "1c. KDE contour",
          subtitle = "kde_threshold=0.02, buffer=0.1",
          fill = "#27ae60", border = "#1e8449") +
plot_annotation(
  title = "Section 1: Circular Cloud — Sanity Check",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**What to look for:** All three masks should closely envelope the point cloud. Small differences near the boundary are expected — convex gives a polygon, KDE gives a smoother contour. Any large discrepancy here indicates a parameter problem.

---

# Section 2 — Elongated cloud: anisotropy

Elongated tissue (e.g. a muscle section, a linear airway epithelium) has very different extents along its two axes. All methods should handle this naturally; this section confirms that no method is implicitly assuming isotropy.

```{r sec2}
coords_elong <- data.frame(x = rnorm(n, 0, 3),
                            y = rnorm(n, 0, 0.6))

m2a <- fit_spatial_mask(coords_elong, method = "convex",  verbose = FALSE)
m2b <- fit_spatial_mask(coords_elong, method = "concave", verbose = FALSE)
m2c <- fit_spatial_mask(coords_elong, method = "kde",
                         kde_threshold = 0.02, buffer_dist = 0.1,
                         verbose = FALSE)
```

```{r sec2-plot, fig.height=4}
plot_mask(m2a, coords_elong, title = "2a. Convex") +
plot_mask(m2b, coords_elong, title = "2b. Concave") +
plot_mask(m2c, coords_elong, title = "2c. KDE",
          fill = "#27ae60", border = "#1e8449") +
plot_annotation(
  title = "Section 2: Elongated Cloud",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**What to look for:** The mask should track the long axis of the cloud, not default to a circle. The KDE bandwidth is computed separately for x and y (`MASS::bandwidth.nrd`), so it adapts to anisotropy automatically.

---

# Section 3 — L-shaped cloud: the non-convex diagnostic

This is the most important basic test. The L-shape is the canonical non-convex case: the convex hull incorrectly fills the concave corner of the L, including empty space that contains no tissue. A good mask must track the actual boundary and exclude the corner.

> **Rule of thumb:** If your tissue has any re-entrant angles (C-shapes, crescents, airways, arcs), do not use `method = "convex"`. Use `method = "concave"` or `method = "raster"`.

```{r sec3-data}
coords_L <- rbind(
  data.frame(x = runif(200, 0, 5) + rnorm(200, 0, 0.05),
             y = runif(200, 0, 1) + rnorm(200, 0, 0.05)),   # horizontal bar
  data.frame(x = runif(200, 0, 1) + rnorm(200, 0, 0.05),
             y = runif(200, 0, 5) + rnorm(200, 0, 0.05)))   # vertical bar
```

```{r sec3-fit}
m3a <- fit_spatial_mask(coords_L, method = "convex",  verbose = FALSE)
m3b <- fit_spatial_mask(coords_L, method = "concave", concavity = 1.5, verbose = FALSE)
m3c <- fit_spatial_mask(coords_L, method = "concave", concavity = 3.0, verbose = FALSE)
```

```{r sec3-plot, fig.height=4}
plot_mask(m3a, coords_L,
          title = "3a. Convex hull",
          subtitle = "Incorrectly includes empty corner",
          fill = "#e74c3c", border = "#922b21") +
plot_mask(m3b, coords_L,
          title = "3b. Concave, tight (1.5)",
          subtitle = "Correctly excludes corner") +
plot_mask(m3c, coords_L,
          title = "3c. Concave, relaxed (3.0)",
          subtitle = "Partially includes corner",
          fill = "#8e44ad", border = "#6c3483") +
plot_annotation(
  title    = "Section 3: L-Shaped Cloud — Non-Convex",
  subtitle = "Lower concavity wraps more tightly; convex hull fails",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

### The `concavity` parameter

`concavity` controls how aggressively the hull "cuts in" to follow the data boundary. It is the key tuning parameter for `method = "concave"`.

| Value | Behaviour |
|---|---|
| 1.0–1.5 | Very tight — tracks the data boundary closely |
| 2.0 | Default — good balance for most tissue shapes |
| 3.0–5.0 | Relaxed — approaches the convex hull |

**When to use lower values:** C-shapes, crescents, U-shapes, any tissue with a notch or bay. **When to use higher values:** Dense blobs where you want a smooth, padded boundary. If you see the mask cutting through the interior of your tissue, increase `concavity`. If it fills in empty space at concavities, decrease it.

---

# Section 4 — Multi-cluster blobs: connected vs. isolated masks

When tissue fragments are well-separated, a low `kde_threshold` will produce a single mask that bridges them (because the KDE density never drops to zero between clusters at low thresholds). A higher threshold separates them into distinct polygons.

The raster method's `raster_sigma` plays the analogous role: a large sigma bridges clusters; a small sigma isolates them.

```{r sec4}
mk_cluster <- function(n, cx, cy, sd = 0.4)
  data.frame(x = rnorm(n, cx, sd), y = rnorm(n, cy, sd))

coords_multi <- rbind(mk_cluster(150, 0, 0),
                       mk_cluster(150, 4, 1),
                       mk_cluster(150, 2, 4))

m4a <- fit_spatial_mask(coords_multi, method = "convex",  verbose = FALSE)
m4b <- fit_spatial_mask(coords_multi, method = "concave", concavity = 2,
                         verbose = FALSE)
m4c <- fit_spatial_mask(coords_multi, method = "kde",
                         kde_threshold = 0.01, buffer_dist = 0.15,
                         verbose = FALSE)
m4d <- fit_spatial_mask(coords_multi, method = "kde",
                         kde_threshold = 0.25, buffer_dist = 0.15,
                         verbose = FALSE)
```

```{r sec4-plot, fig.height=7}
(plot_mask(m4a, coords_multi, title = "4a. Convex") +
 plot_mask(m4b, coords_multi, title = "4b. Concave")) /
(plot_mask(m4c, coords_multi,
           title    = "4c. KDE — inclusive",
           subtitle = "threshold=0.01: merged",
           fill = "#27ae60", border = "#1e8449") +
 plot_mask(m4d, coords_multi,
           title    = "4d. KDE — tight",
           subtitle = "threshold=0.25: separated",
           fill = "#8e44ad", border = "#6c3483")) +
plot_annotation(
  title = "Section 4: Multi-Cluster",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

### Choosing `kde_threshold`

`kde_threshold` is expressed as a quantile of the KDE density values across the grid.

- Low values (0.01–0.05): the contour is drawn far out in the low-density tails. Clusters merge; mask is generous.
- High values (0.15–0.30): only the dense core is enclosed. Clusters separate; mask is tight.

**Practical advice:** Start at 0.05 and increase until the mask matches your visual inspection. If you expect biologically separated regions (e.g. two lobes), a threshold that separates them is usually correct.

---

# Section 5 — Concavity sweep on a crescent

A crescent is a severe test of the concave hull: it has a large, deep concavity spanning most of the shape. This sweep lets you calibrate the `concavity` parameter on a shape where the expected correct answer is obvious.

```{r sec5}
theta        <- seq(0.2, pi - 0.2, length.out = 300)
coords_cres  <- data.frame(
  x = cos(theta) * (1 + rnorm(300, 0, 0.08)) * 3,
  y = sin(theta) * (1 + rnorm(300, 0, 0.08)) * 3 - cos(theta) * 1.5)

pal <- c("#e74c3c", "#e67e22", "#27ae60", "#2980b9")
sw  <- mapply(function(cv, col) {
  m <- fit_spatial_mask(coords_cres, method = "concave",
                         concavity = cv, verbose = FALSE)
  plot_mask(m, coords_cres, title = paste0("concavity = ", cv),
            fill = col, border = col)
}, c(1.2, 2, 3, 5), pal, SIMPLIFY = FALSE)
```

```{r sec5-plot, fig.height=4}
wrap_plots(sw, nrow = 1) +
  plot_annotation(
    title    = "Section 5: Concavity Sweep — Crescent",
    subtitle = "Lower = tighter | Higher = approaches convex hull",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 9, color = "grey40")))
```

**What to look for:** At `concavity = 1.2`, the hull cuts tightly around the arc. At `concavity = 5`, it is nearly convex and fills the interior. The "correct" value for real tissue is the one that matches a human drawing of the tissue boundary. For a crescent-shaped airway section, `1.2–1.5` is usually appropriate.

---

# Section 6 — Buffer and smoothing

After fitting, two post-processing options are available:

- **`buffer_dist`**: expands the mask outward by this distance in coordinate units using `sf::st_buffer()`. Useful when you want a small safety margin to ensure all cells fall inside the mask, or when points are sparse near the boundary.
- **`smooth_mask`**: applies `sf::st_simplify()` to reduce the vertex count and round jagged boundary segments. Controlled by `smooth_tolerance` (default: 1% of bounding box diagonal).

```{r sec6}
r           <- sqrt(runif(250, 0.4^2, 1^2))
ang         <- runif(250, 0, 2 * pi)
coords_ring <- data.frame(x = r * cos(ang) * 5,
                           y = r * sin(ang) * 5)

m6a <- fit_spatial_mask(coords_ring, method = "concave", concavity = 2,
                         buffer_dist = 0,   smooth_mask = FALSE, verbose = FALSE)
m6b <- fit_spatial_mask(coords_ring, method = "concave", concavity = 2,
                         buffer_dist = 0.3, smooth_mask = FALSE, verbose = FALSE)
m6c <- fit_spatial_mask(coords_ring, method = "concave", concavity = 2,
                         buffer_dist = 0.3, smooth_mask = TRUE,
                         smooth_tolerance = 0.15, verbose = FALSE)
```

```{r sec6-plot, fig.height=4}
plot_mask(m6a, coords_ring,
          title = "6a. Raw — no buffer, no smooth") +
plot_mask(m6b, coords_ring,
          title = "6b. Buffer = 0.3",
          fill = "#27ae60", border = "#1e8449") +
plot_mask(m6c, coords_ring,
          title = "6c. Buffer + smooth",
          subtitle = "smooth_tolerance = 0.15",
          fill = "#8e44ad", border = "#6c3483") +
plot_annotation(
  title = "Section 6: Buffer and Smoothing",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

**When to use each:**

- `buffer_dist`: use when downstream point-in-mask tests are failing for points you know should be inside (e.g. points right on the boundary). A value of 1–5% of the domain width is usually safe.
- `smooth_mask`: use when you want a visually clean boundary for publication figures, or when a jagged boundary is causing downstream geometric operations to be slow or numerically unstable. Be careful: large `smooth_tolerance` values can erase real tissue features.

---

# Section 7 — Topologically complex cases: holes and islands

The `"raster"` method is the recommended approach for masks that need to represent **holes** (vessel lumens, necrotic cores, tissue tears) or **disconnected islands**. It is also the fastest method for large point clouds (100k+ cells).

## How the raster method works

1. **Bin** all points onto a regular `raster_resolution × raster_resolution` grid.
2. **Convolve** the binary occupancy grid with a 2D isotropic Gaussian of width `raster_sigma` (in coordinate units). This is equivalent to placing a Gaussian "blob" at every point and summing — the value at each cell represents how much point density is nearby.
3. **Threshold** at `raster_threshold × max(field)`. Cells above threshold are "inside"; cells below are "outside".
4. **Build** one axis-aligned rectangle per inside cell, then **dissolve** all rectangles with `sf::st_union()`. GEOS handles the topology automatically — holes and islands emerge from the geometry without any ring-winding logic.
5. **Smooth** the staircase boundary with a morphological close (buffer out then in by ~half the cell diagonal).

## Key tuning parameters

| Parameter | Effect |
|---|---|
| `raster_sigma` ↑ | Holes fill in, islands merge, boundary smooths |
| `raster_sigma` ↓ | Fine holes and gaps preserved |
| `raster_threshold` ↑ | Mask shrinks; requires denser local coverage |
| `raster_threshold` ↓ | Mask grows; accepts sparse regions |
| `raster_resolution` ↑ | Finer grid; more detail; slower union step |

> **Units note:** `raster_sigma` is in **coordinate units** (same as x and y). If your tissue coordinates span 0–10, a `raster_sigma` of 0.25 means each Gaussian blob has a standard deviation of 0.25 tissue units. `NULL` sets sigma automatically to 3% of the domain width.

---

## 7a — Donut: single interior void

```{r sec7a}
r_d          <- sqrt(runif(800, 2^2, 5^2))
th_d         <- runif(800, 0, 2 * pi)
coords_donut <- data.frame(x = r_d * cos(th_d),
                            y = r_d * sin(th_d))

m7a_r <- fit_spatial_mask(coords_donut, method = "raster",
  raster_resolution = 256L, raster_sigma = 0.25, raster_threshold = 0.2,
  verbose = FALSE)
m7a_c <- fit_spatial_mask(coords_donut, method = "convex", verbose = FALSE)
```

```{r sec7a-plot, fig.height=4}
plot_mask(m7a_r, coords_donut,
          title    = "7a-i. Raster — hole detected ✓",
          subtitle = "σ = 0.25 coord units, threshold = 0.2") +
plot_mask(m7a_c, coords_donut,
          title    = "7a-ii. Convex — hole filled ✗",
          subtitle = "Convex hull cannot represent holes",
          fill = "#e74c3c", border = "#922b21") +
plot_annotation(
  title = "Donut: Single Interior Void",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

The raster method naturally detects the empty centre as a region of low convolved density, below the threshold. The convex hull cannot ever represent a hole — it is by definition the smallest convex polygon containing all the points.

---

## 7b — Swiss cheese: multiple holes

```{r sec7b}
.circle_mat <- function(cx, cy, r, n = 100) {
  th  <- seq(0, 2 * pi, length.out = n)
  mat <- cbind(cx + r * cos(th), cy + r * sin(th))
  rbind(mat, mat[1L, ])
}

hdefs <- list(
  list(cx = 2.5, cy = 7.5, r = 1.1),
  list(cx = 7.0, cy = 6.5, r = 1.4),
  list(cx = 4.0, cy = 2.5, r = 0.9),
  list(cx = 8.0, cy = 2.0, r = 1.2)
)

sample_in_polygon <- function(poly_sf, n, max_tries = 20) {
  bb <- sf::st_bbox(poly_sf); mu <- sf::st_union(poly_sf)
  pts <- data.frame(x = numeric(0), y = numeric(0))
  for (i in seq_len(max_tries)) {
    if (nrow(pts) >= n) break
    cnd <- data.frame(x = runif(n * 8, bb["xmin"], bb["xmax"]),
                      y = runif(n * 8, bb["ymin"], bb["ymax"]))
    ok  <- as.logical(sf::st_within(
      sf::st_as_sf(cnd, coords = c("x","y")), mu, sparse = FALSE)[, 1L])
    pts <- rbind(pts, cnd[ok, ])
  }
  pts[seq_len(min(n, nrow(pts))), ]
}

outer_sq     <- sf::st_sfc(sf::st_polygon(list(
  rbind(c(0,0), c(10,0), c(10,10), c(0,10), c(0,0)))))
hcircles     <- lapply(hdefs, function(h)
  sf::st_polygon(list(.circle_mat(h$cx, h$cy, h$r))))
swiss_region <- sf::st_difference(outer_sq,
  sf::st_sfc(sf::st_union(sf::st_sfc(hcircles))))
coords_swiss <- sample_in_polygon(swiss_region, n = 1500)

m7b_r <- fit_spatial_mask(coords_swiss, method = "raster",
  raster_resolution = 256L, raster_sigma = 0.25, raster_threshold = 0.2,
  verbose = FALSE)
m7b_c <- fit_spatial_mask(coords_swiss, method = "concave",
  concavity = 1.8, verbose = FALSE)
```

```{r sec7b-plot, fig.height=4}
plot_mask(m7b_r, coords_swiss,
          title    = "7b-i. Raster — all 4 holes detected ✓",
          subtitle = "σ = 0.25, threshold = 0.2") +
plot_mask(m7b_c, coords_swiss,
          title    = "7b-ii. Concave — holes filled ✗",
          subtitle = "Concave hull also cannot represent holes",
          fill = "#e74c3c", border = "#922b21") +
plot_annotation(
  title = "Swiss Cheese: Multiple Interior Holes",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

---

## 7c — Archipelago: disconnected islands

```{r sec7c}
mk_isl <- function(n, cx, cy, rx, ry, ns = 0.15) {
  th <- runif(n, 0, 2 * pi); r <- sqrt(runif(n))
  data.frame(x = cx + r * rx * cos(th) + rnorm(n, 0, ns),
             y = cy + r * ry * sin(th) + rnorm(n, 0, ns))
}
coords_arch <- rbind(mk_isl(300,  0,  0, 3,   2),
                      mk_isl(300,  9,  4, 2,   3),
                      mk_isl(300,  5, -7, 2.5, 1.5))

m7c_r <- fit_spatial_mask(coords_arch, method = "raster",
  raster_resolution = 256L, raster_sigma = 0.4, raster_threshold = 0.18,
  verbose = FALSE)
m7c_c <- fit_spatial_mask(coords_arch, method = "convex", verbose = FALSE)

n_geom <- length(sf::st_cast(m7c_r, "POLYGON"))
```

```{r sec7c-plot, fig.height=4}
plot_mask(m7c_r, coords_arch,
          title    = sprintf("7c-i. Raster — %d separate polygons ✓", n_geom),
          subtitle = "Each island detected independently") +
plot_mask(m7c_c, coords_arch,
          title    = "7c-ii. Convex — islands bridged ✗",
          subtitle  = "Convex hull merges all islands into one",
          fill = "#e74c3c", border = "#922b21") +
plot_annotation(
  title = "Archipelago: Disconnected Islands",
  theme = theme(plot.title = element_text(face = "bold", size = 13)))
```

Disconnected islands emerge automatically from the GEOS union step: if two groups of "on" cells are not touching, they become separate polygons within the same `sfc` object. The number of distinct polygons is accessible via `length(sf::st_cast(mask, "POLYGON"))`.

---

## 7d — Sigma sensitivity sweep

This is the most important tuning exercise. Run it on your own data before finalising `raster_sigma`.

```{r sec7d}
sig_vals <- c(0.1, 0.2, 0.4, 0.7)
sig_cols <- c("#e74c3c", "#e67e22", "#27ae60", "#2980b9")

sig_pls <- mapply(function(sv, col) {
  m  <- fit_spatial_mask(coords_swiss, method = "raster",
    raster_resolution = 256L, raster_sigma = sv,
    raster_threshold  = 0.2,  verbose = FALSE)
  nh <- max(0L, length(sf::st_cast(m, "POLYGON")) - 1L)
  plot_mask(m, coords_swiss,
            title    = paste0("σ = ", sv, " coord units"),
            subtitle = paste0(nh, " hole(s) detected"),
            fill = col, border = col, pt_size = 0.5)
}, sig_vals, sig_cols, SIMPLIFY = FALSE)
```

```{r sec7d-plot, fig.height=4}
wrap_plots(sig_pls, nrow = 1) +
  plot_annotation(
    title    = "7d: raster_sigma Sensitivity — Swiss Cheese",
    subtitle = "Small σ → holes preserved | Large σ → holes filled",
    theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                     plot.subtitle = element_text(size = 9, color = "grey40")))
```

**Tuning strategy:**

1. Start with `raster_sigma = NULL` (auto = 3% of domain width)
2. Inspect the result; note whether real voids are detected
3. Decrease sigma if holes are being filled that you expect to be present
4. Increase sigma if the mask is too jagged or breaks into spurious fragments where there is actually continuous tissue
5. Adjust `raster_threshold` to expand or contract the mask after sigma is set

---

## 7e — Nested crescent voids

```{r sec7e}
mk_cres <- function(cx, cy, ro, ri, dx, dy, n = 100) {
  moc <- .circle_mat(cx,    cy,    ro, n)
  mic <- .circle_mat(cx+dx, cy+dy, ri, n)
  oc  <- sf::st_polygon(list(moc))
  ic  <- sf::st_polygon(list(mic))
  sf::st_sfc(sf::st_difference(sf::st_sfc(oc), sf::st_sfc(ic)))
}

th_od  <- seq(0, 2 * pi, length.out = 200)
mat_od <- cbind(10 * cos(th_od), 10 * sin(th_od))
mat_od <- rbind(mat_od, mat_od[1L, ])
outer_d       <- sf::st_sfc(sf::st_polygon(list(mat_od)))
cres1         <- mk_cres(-2, 2,  2.5, 2,   0.9, -0.2)
cres2         <- mk_cres( 3, -3, 2.2, 1.8, -0.7,  0.4)
nest_region   <- sf::st_difference(outer_d, sf::st_union(c(cres1, cres2)))
coords_nested <- sample_in_polygon(nest_region, n = 2500)

m7e_fine   <- fit_spatial_mask(coords_nested, method = "raster",
  raster_resolution = 256L, raster_sigma = 0.4,  raster_threshold = 0.18,
  verbose = FALSE)
m7e_coarse <- fit_spatial_mask(coords_nested, method = "raster",
  raster_resolution = 256L, raster_sigma = 1.0, raster_threshold = 0.18,
  verbose = FALSE)
```

```{r sec7e-plot, fig.height=4}
plot_mask(m7e_fine,   coords_nested,
          title    = "7e-i. Fine σ = 0.4",
          subtitle = "Crescent voids preserved") +
plot_mask(m7e_coarse, coords_nested,
          title    = "7e-ii. Coarse σ = 1.0",
          subtitle = "Narrow voids filled in",
          fill = "#8e44ad", border = "#6c3483") +
plot_annotation(
  title    = "Nested Crescent Voids: sigma controls void resolution",
  subtitle = "The minimum detectable void width is approximately 2 × raster_sigma",
  theme    = theme(plot.title    = element_text(face = "bold", size = 13),
                   plot.subtitle = element_text(size = 9, color = "grey40")))
```

> **Key principle:** The minimum resolvable void has a width of approximately `2 × raster_sigma`. Voids narrower than this will be filled in by the Gaussian blur. If you need to resolve fine voids (e.g. individual capillary lumens), you must use a small sigma — but this may also make the mask jagged elsewhere.

---

## 7f — Scale test: 100,000 cells

The raster method is `O(N²)` in the grid size but `O(n)` in the number of input points (binning is a single `tabulate()` call). This makes it practical for large datasets.

```{r sec7f, cache=TRUE}
th100    <- seq(0, 2 * pi, length.out = 400)
r100     <- 12 + sin(5 * th100) + 0.4 * cos(11 * th100)
mat100   <- cbind(r100 * cos(th100), r100 * sin(th100))
mat100   <- rbind(mat100, mat100[1L, ])
outer100 <- sf::st_sfc(sf::st_polygon(list(mat100)))

make_ellipse_poly <- function(cx, cy, a, b, angle_deg, n = 80) {
  th  <- seq(0, 2 * pi, length.out = n)
  ang <- angle_deg * pi / 180
  xe  <- a * cos(th); ye <- b * sin(th)
  mat <- cbind(cx + xe * cos(ang) - ye * sin(ang),
               cy + xe * sin(ang) + ye * cos(ang))
  mat <- rbind(mat, mat[1L, ])
  sf::st_polygon(list(mat))
}

holes100 <- sf::st_sfc(list(
  make_ellipse_poly( 3,  5,  3,   1.5,  10),
  make_ellipse_poly(-5,  2,  2.5, 2,   -30),
  make_ellipse_poly( 1, -6,  2,   3,    50)
))
reg100      <- sf::st_difference(outer100, sf::st_union(holes100))
coords_100k <- sample_in_polygon(reg100, n = 100000)

t_raster <- system.time(
  m_100k <- fit_spatial_mask(coords_100k, method = "raster",
    raster_resolution = 256L, raster_sigma = 0.6,
    raster_threshold  = 0.18, verbose = TRUE)
)
```

```{r sec7f-plot, fig.height=5}
sub <- coords_100k[sample(nrow(coords_100k), 5000), ]
plot_mask(m_100k, sub,
          title    = sprintf("7f. 100k cells — raster (%.1f s)", t_raster["elapsed"]),
          subtitle = "3 elliptic holes detected | 5,000 of 100,000 points shown",
          pt_size  = 0.3, pt_alpha = 0.2)
```

---

# Section 8 — Validation: point containment

A properly fitted mask must contain all its input points. This table checks every mask produced in this vignette. Any FAIL indicates that the mask needs a larger `buffer_dist` or a lower `raster_threshold`.

```{r sec8-validation}
cases <- list(
  list(label = "Circle/convex",     mask = m1a, coords = coords_circle),
  list(label = "Circle/concave",    mask = m1b, coords = coords_circle),
  list(label = "Circle/kde",        mask = m1c, coords = coords_circle),
  list(label = "Elongated/convex",  mask = m2a, coords = coords_elong),
  list(label = "Elongated/concave", mask = m2b, coords = coords_elong),
  list(label = "Elongated/kde",     mask = m2c, coords = coords_elong),
  list(label = "L-shape/convex",    mask = m3a, coords = coords_L),
  list(label = "L-shape/conc1.5",   mask = m3b, coords = coords_L),
  list(label = "L-shape/conc3",     mask = m3c, coords = coords_L),
  list(label = "Multi/convex",      mask = m4a, coords = coords_multi),
  list(label = "Multi/concave",     mask = m4b, coords = coords_multi),
  list(label = "Multi/kde-lo",      mask = m4c, coords = coords_multi),
  list(label = "Multi/kde-hi",      mask = m4d, coords = coords_multi),
  list(label = "Ring/raw",          mask = m6a, coords = coords_ring),
  list(label = "Ring/buffer",       mask = m6b, coords = coords_ring),
  list(label = "Ring/smooth",       mask = m6c, coords = coords_ring),
  list(label = "Donut/raster",      mask = m7a_r, coords = coords_donut),
  list(label = "Swiss/raster",      mask = m7b_r, coords = coords_swiss),
  list(label = "Archipelago/raster",mask = m7c_r, coords = coords_arch),
  list(label = "Nested/fine",       mask = m7e_fine,   coords = coords_nested),
  list(label = "Nested/coarse",     mask = m7e_coarse, coords = coords_nested),
  list(label = "100k/raster",       mask = m_100k,     coords = coords_100k)
)

results <- do.call(rbind, lapply(cases, function(vc) {
  pts <- sf::st_geometry(sf::st_as_sf(vc$coords, coords = c("x","y")))
  ok  <- sf::st_within(pts, sf::st_union(vc$mask), sparse = FALSE)[, 1L]
  n   <- nrow(vc$coords)
  data.frame(
    Case     = vc$label,
    N        = n,
    Enclosed = sum(ok),
    Pct      = round(100 * sum(ok) / n, 2),
    Result   = ifelse(sum(ok) == n, "✓ PASS", "✗ FAIL"),
    stringsAsFactors = FALSE
  )
}))

results
```

---

# Quick-start decision guide

```
Your data
  │
  ├─ Simple convex blob? ──────────────────────► method = "convex"
  │
  ├─ Non-convex (C-shape, arc, L-shape)?
  │   └─ No holes needed? ───────────────────► method = "concave"
  │                                             concavity: start at 2, lower to tighten
  │
  ├─ Holes or islands expected?
  │   ├─ n < 50,000? ─────────────────────────► method = "raster"
  │   └─ n > 50,000? ─────────────────────────► method = "raster" (still fine)
  │
  └─ Dense, smooth, no need for holes?
      └─ KDE is an alternative ───────────────► method = "kde"
                                                kde_threshold: start at 0.05

For method = "raster":
  raster_sigma   = NULL  (auto) for a first pass
  raster_sigma ↓ to preserve fine voids
  raster_sigma ↑ to fill gaps and smooth edges
  raster_threshold = 0.15 is a reliable default
```

---

# Session information

```{r session-info}
sessionInfo()
```
